---
title: "Correct covariance handling in the rpact–emmeans interface"
date: 2026-09-11
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  message: false
params:
  vignette_csv: null
---

### Scope

- Two-arm adjusted continuous-outcome analyses, with required extensions for multi-arm comparisons and longitudinal mixed models from the existing development vignettes.
- Option B is the implementation target: preserve model means and joint contrasts for ordinary Gaussian linear models and the longitudinal `lmer` models in both development vignettes, fitted separately to independent subject cohorts.

### Source baseline

- rpact Git commit `8c9302e4`, package version `4.5.0.9319`.
- The execution environment used for this render is recorded below; the commit above identifies the original investigation baseline.

## Rendering and reproducibility

Render from the repository root with
`quarto render misc/emmeans-interface-design.qmd`. This document loads the local
`rpact` checkout and requires `pkgload`, `emmeans`, `lme4`, `pbkrtest`, and `mvtnorm`.
All example and verification code is included below.

The vignette CSV is external to this repository. Set the environment variable
`RPACT_VIGNETTE_DATA` to its path, or pass
`-P vignette_csv:/path/to/data_two_arm_adjusted_means.csv` when rendering.
The default also checks the original SVN location under the current user's home.
If the adjusted-means CSV is unavailable, its reproduction is skipped with a
visible note. The longitudinal examples require both CSVs in `RPACT_VIGNETTE_DATA_DIR`
(default: the SVN checkout's `inst/data` directory under the current user's home);
missing files or mixed-model dependencies produce an actionable error.

```{r}
#| label: setup
repo_root <- normalizePath("..", mustWork = TRUE)
stopifnot(file.exists(file.path(repo_root, "DESCRIPTION")))
suppressPackageStartupMessages(library(emmeans))
document_dir <- getwd()
pkgload::load_all(repo_root, quiet = TRUE)
setwd(document_dir)
for (package in c("lme4", "pbkrtest", "mvtnorm")) {
    if (!requireNamespace(package, quietly = TRUE)) {
        stop("Install package '", package, "' to run the longitudinal prototypes.")
    }
}
vignette_data_dir <- Sys.getenv("RPACT_VIGNETTE_DATA_DIR", unset = path.expand(
    "~/svn/rpact/rpact-vignettes/inst/data"))
for (filename in c("data_multi_arm_means_repeated_measures.csv",
                   "data_two_arm_means_repeated_measures.csv")) {
    if (!file.exists(file.path(vignette_data_dir, filename))) {
        stop("Set RPACT_VIGNETTE_DATA_DIR to the SVN inst/data directory: missing ", filename)
    }
}
options(rpact.dataset.stdev.calc.mode = "auto")
Sys.unsetenv("RPACT_DATASET_STDEV_CALC_MODE")
if (!exists("params")|| is.null(vignette_csv <- params$vignette_csv)) {
    vignette_csv <- Sys.getenv("RPACT_VIGNETTE_DATA", unset = path.expand(
        "~/svn/rpact/rpact-vignettes/inst/exclusive/qmd/data/data_two_arm_adjusted_means.csv"
    ))
}
```

This render uses R `r getRversion()` and `emmeans` `r packageVersion("emmeans")`.

The following helpers compare the legacy adapter with direct `emmeans` inference
and the existing general-estimates interface. Their assertions check the
contrast variance and agreement of the corrected stage tests.

```{r}
#| label: comparison-helpers

# emm: emmeans object representing the estimated marginal means for the treatment and control groups.
# treatment: name of the treatment group.
# control: name of the control group.
# Returns a data frame containing the contrast estimate, its standard error, and associated statistics.
extract_contrast <- function(emm, treatment, control) {
    tab <- as.data.frame(summary(emm, type = "link"))
    stopifnot(nrow(tab) == 2L,
        setequal(as.character(tab$group), c(control, treatment)))
    coefficients <- ifelse(tab$group == treatment, 1, -1)
    effect <- contrast(emm, list("Treatment - Control" = coefficients), adjust = "none")
    result <- as.data.frame(summary(effect, type = "link", infer = c(TRUE, TRUE),
        null = 0, side = "two-sided", adjust = "none"))
    manual_se <- sqrt(drop(t(coefficients) %*% vcov(emm) %*% coefficients))
    stopifnot(isTRUE(all.equal(manual_se, result$SE, tolerance = 1e-12)))
    result
}

# Returns a data frame containing the comparison of the legacy adapter, general estimates interface, and direct `emmeans` inference for the specified contrast.
# Checks corrected general-estimates inference against emmeans; the legacy result may differ.
compare_paths <- function(emm, treatment, control) {
    contrast_result <- extract_contrast(emm, treatment, control)
    design <- getDesignGroupSequential(kMax = 1)
    legacy <- getStageResults(design,
        dataInput = suppressWarnings(getDataset(emm)))
    general <- getStageResults(design, dataInput = getDataset(
        est = contrast_result$estimate, se = contrast_result$SE, df = contrast_result$df
    ))
    correct_p <- pt(contrast_result$estimate / contrast_result$SE,
        df = contrast_result$df, lower.tail = FALSE)
    emm_table <- as.data.frame(summary(emm, type = "link"))
    effect <- contrast(emm, list("Treatment - Control" =
        ifelse(emm_table$group == treatment, 1, -1)), adjust = "none")
    direct_test <- as.data.frame(emmeans::test(effect,
        null = 0, side = ">", adjust = "none"))
    stopifnot(
        isTRUE(all.equal(direct_test$p.value, correct_p, tolerance = 1e-12)),
        isTRUE(all.equal(general$testStatistics[1],
            contrast_result$estimate / contrast_result$SE, tolerance = 1e-12)),
        isTRUE(all.equal(general$pValues[1], correct_p, tolerance = 1e-12))
    )
    data.frame(
        estimate = contrast_result$estimate,
        covariance = vcov(emm)[1, 2],
        diagonal_only_se = sqrt(sum(diag(vcov(emm)))),
        legacy_se = contrast_result$estimate / legacy$testStatistics[1],
        correct_se = contrast_result$SE,
        contrast_df = contrast_result$df,
        legacy_p = legacy$pValues[1],
        correct_p = correct_p
    )
}
```

## 1. Current interface and its assumptions

### Entry points and conversion

The vignette
`rpact-vignettes/inst/exclusive/qmd/rpact_analysis_two_arm_adjusted_means_example.qmd`
fits `lm(outcome ~ group + covariate + gender)` independently within each stage,
calls `emmeans(model, "group")`, and passes one or several resulting objects to
`getDataset()`. Thus stage 2 means new stage-2 observations, not all observations
available at the second look.

The relevant implementation is:

| Component | Current responsibility |
|---|---|
| [`getDataset()` / `.getDataset()`](../R/class_analysis_dataset.R) | Accept an `emmGrid`, several such arguments, or a single list of stage objects; dispatch before ordinary numeric dataset parsing. |
| `.getDatasetMeansModelObjectsList()` in the same file | Collect `emmGrid` arguments; warn and ignore other arguments when EMM objects are present. |
| `.getDatasetMeansFromModelsByStage()` in the same file | Extract per-group summaries, construct synthetic means data, then call `getDataset()` again on a wide data frame. |
| `.getStandardDeviationFromStandardError()` in the same file | Convert marginal SEs into quantities stored as group SDs. |
| `DatasetMeans` in the same file | R6 subclass of `Dataset`, storing stage/group means, SDs and sample sizes, and their cumulative versions. |
| [Means analysis](../R/f_analysis_base_means.R) | Apply ordinary independent-sample formulas, producing `StageResultsMeans` and design-specific analysis results. |

An `emmGrid` is an `emmeans` S4 object. It retains the linear functions of model
coefficients defining the EMMs, coefficient estimates and covariance, and a
degrees-of-freedom calculation mechanism. `summary()` exposes marginal EMMs,
SEs and df; `vcov()` exposes the covariance matrix of the EMM estimates on the
linear-predictor scale. Covariance is available without refitting the model.
See the [emmeans object methods](https://rvlenth.github.io/emmeans/reference/emmGrid-methods.html).

The adapter currently:

1. Requires `emmGrid` stage objects. It rejects objects with more than one entry
   in their `levels` slot. The error says that models with covariates are not
   supported, but the actual check concerns the *returned grid*: an adjusted
   model summarized over `group` alone passes, as in the vignette.
2. Determines whether the model call looks like `lm` from the first stage only.
   It warns about imprecision for `lm` and experimental support for other models.
3. Reads `emmean`, `SE`, `df` from `summary(object)`, and `sigma` and `extras`
   from `as.list(object)`. Rows of `extras` are treated as sample sizes. These
   reference-grid weights are not a general contract for independent subject
   counts, particularly for weighted or repeated-measures models.
4. Moves the first EMM level to the last `rpact` group position. For two arms,
   EMM row 2 becomes `rpact` treatment group 1 and row 1 becomes control group 2.
   This assumes that the first EMM level is the control. It does not identify
   treatment from a scientific label. Choosing the control as the regression
   reference is a convenience for this convention, not a statistical requirement
   for calculating a treatment-minus-control contrast.
5. Under the default `auto` mode for `lm`, sets
   `stDev[g] = SE(EMM[g]) * sqrt(n[g])`. The environment variable
   `RPACT_DATASET_STDEV_CALC_MODE` / option `rpact.dataset.stdev.calc.mode`
   also permits `sigma`, `norm`, and `t` modes. `sigma` uses residual model SD;
   `t` uses `SE * 2 / qt(0.975, df) * sqrt(n)`. For non-`lm` models with finite
   positive df, the latter conversion also applies outside `sigma` mode.
6. Returns `DatasetMeans`, retaining neither EMM covariance nor model/contrast
   df. There is no model-aware dataset class in this conversion path.

### Downstream assumptions

For a two-arm dataset, the default is `equalVariances = TRUE` and
`normalApproximation = FALSE`. At a single stage, the means path calculates

$$
s_p^2=\frac{(n_T-1)s_T^2+(n_C-1)s_C^2}{n_T+n_C-2},\qquad
SE_{\mathrm{pooled}}=s_p\sqrt{1/n_T+1/n_C},
$$

and uses `n_T + n_C - 2` df. Setting `equalVariances = FALSE` instead uses
`sqrt(s_T^2/n_T + s_C^2/n_C)` with Welch df. Neither path includes covariance.
With synthetic SDs and unequal group sizes, even the pooled SE need not equal
the square root of the sum of the marginal EMM variances.

Cumulative means and SDs are constructed as ordinary sample summaries. This
does not generally reconstruct either a pooled ANCOVA fit or an
inverse-variance combination of independently estimated treatment contrasts.

The reduction is appropriate for a conventional independent, homoscedastic
Gaussian `lm(y ~ group)`: the EMMs are the group sample means, their covariance
is zero, and the model df are `n_T + n_C - 2`.

However, covariate adjustment invalidates
this reduction in general. Additional covariates do not *always* cause nonzero
covariance: balanced/orthogonal designs can still give zero covariance. Shared
coefficients alone are not a proof of correlation—the full covariance
calculation decides. On the other hand, even a zero covariance does not justify discarding the model degrees of freedom.

## 2. The statistical problem

Let the fitted coefficient vector be $\hat\beta$, and let the two rows of
$L$ define the requested EMMs. Conditional on the model design,

$$
\hat\mu=L\hat\beta,\qquad V=L\widehat{\operatorname{Var}}(\hat\beta)L^\top.
$$

In EMM order `(Control, Treatment)`, set $c=(-1,1)^\top$. The desired
treatment effect and its variance are

$$
\hat\theta=c^\top\hat\mu,\qquad
\widehat{\operatorname{Var}}(\hat\theta)=c^\top Vc
=V_{TT}+V_{CC}-2V_{TC}.
$$

Keeping marginal SEs alone loses $V_{TC}$. Positive covariance makes the
diagonal-only SE too large; negative covariance makes it too small. In ANCOVA,
adjusting imbalanced groups to a common covariate value can produce negative
covariance. The problem is therefore not necessarily conservative.

There is also an independent degrees of freedom (df) problem: a full-rank ordinary linear model
uses `n - rank(X)` residual df. Mixed models may have contrast-specific df.
These cannot in general be recovered from group counts or the marginal EMM df.
Rescaling an SE by a selected t quantile does not preserve the original test
distribution across null values and significance levels.

### A small reproducible ANCOVA example

This deterministic example has six observations per arm and covariate
imbalance. Its residual vector is orthogonal to all model columns, making the
calculation especially easy to inspect.

```{r}
#| label: toy-ancova
library(emmeans)
d <- data.frame(
    group = factor(rep(c("Control", "Treatment"), each = 6),
                   levels = c("Control", "Treatment")),
    x = c(0:5, 3:8)
)
d$y <- 10 + 2 * (d$group == "Treatment") + 3 * d$x +
    rep(c(1, -1, 0, 0, -1, 1), 2)
fit <- lm(y ~ group + x, data = d)
emm <- emmeans(fit, ~ group)
summary(emm)
```

At the common covariate value `r mean(d$x)`, the control and treatment EMMs are `r summary(emm)$emmean[1]` and `r summary(emm)$emmean[2]`.
Both marginal SEs are `r summary(emm)$SE[1]`, and model df are `r summary(emm)$df[1]`. The EMM covariance is

```{r}
vcov(emm)
```

Thus the correct effect is `r diff(summary(emm)$emmean)` and its SE is

```{r}
sqrt(vcov(emm)["Treatment","Treatment"] + vcov(emm)["Control","Control"] - 2 * vcov(emm)["Treatment","Control"])
```

The diagonal-only SE is

```{r}
sqrt(vcov(emm)["Treatment","Treatment"] + vcov(emm)["Control","Control"])
```

Because the group sizes are equal, this is
also the current adapter's default pooled SE in this example.

```{r}
#| label: toy-comparison
knitr::kable(compare_paths(emm, "Treatment", "Control"), digits = 6)
```

Both the missing covariance and the wrong df affect inference. Simple solutions don't work here:

- Changing `equalVariances` does not fix this balanced-size example.
- Changing to normal approximation does not restore covariance either.
- Using residual `sigma` also fails: it omits the additional variance from covariate imbalance.

### The existing vignette is affected

Using the QMD's own `data/data_two_arm_adjusted_means.csv`, its model formula,
and its control-first level order gives the following *stage-wise*, upper-tail
tests against zero. These are not sequentially adjusted p-values.

```{r}
#| label: vignette-comparison
#| results: asis
if (file.exists(vignette_csv)) {
    raw <- read.csv(vignette_csv)
    stage_table <- do.call(rbind, lapply(sort(unique(raw$stage)), function(k) {
        stage_data <- raw[raw$stage == k, ]
        stage_data$group <- relevel(factor(stage_data$group), "Control group")
        model <- lm(outcome ~ group + covariate + gender, data = stage_data)
        stage_emm <- emmeans(model, "group")
        cbind(stage = k, n = nobs(model),
            compare_paths(stage_emm, "Treatment group", "Control group"))
    }))
    print(knitr::kable(stage_table, digits = 6, row.names = FALSE))
} else {
    cat("Vignette reproduction skipped: CSV not found. Set `vignette_csv` or ",
        "`RPACT_VIGNETTE_DATA` to reproduce this comparison.\n")
}
```


The diagonal-only SEs are `r if (exists("stage_table")) paste(formatC(stage_table$diagonal_only_se, digits = 6, format = "f"), collapse = ", ") else "unavailable (CSV missing)"`;
the small further difference from the current SEs is due to pooling synthetic
SDs with unequal arm sizes.

The larger discrepancy is the omitted covariance.
The vignette's stated stopping decisions and intervals must be regenerated
after choosing the sequential analysis specification; this table alone does
not determine those decisions.

### Additional development vignettes: required coverage

The following sources were reviewed in the `rpact-vignettes` SVN checkout,
under `inst/exclusive/dev/`:

- `rpact_multi_arm_analysis_adjusted_means_example.Rmd`, using
  `inst/data/data_multi_arm_means_repeated_measures.csv`.
- `rpact_analysis_two_arm_means_repeated_measures_example.Rmd`, using
  `inst/data/data_two_arm_means_repeated_measures.csv`.

These are existing workflows to migrate, not evidence that the current adapter
provides valid model-based sequential inference. The examples below are
executable extractions using those CSVs. Section 5 extends them into a joint
dataset and closed-testing prototype, with its analysis assumptions stated explicitly.

#### Multi-arm adjusted means with repeated measures

Despite its filename, the multi-arm example is longitudinal. It fits
`outcome ~ group + gender + covariate * visit + (1 + visit | subject)`
separately at each of three stages, obtains `emmeans(model, "group")`, and
passes successive stage grids to `getDataset()` for `getDesignInverseNormal()`
and `getAnalysisResults()`. Its two treatment groups share a control. The
random-effects term permits intercept–slope covariance, despite the source's
wording "without interaction between random effects." The final bulk-import
examples additionally cover group-only `lm`, adjusted `lm`, and a random-intercept
model with `covariate * gender`.

The corrected interface must form both named treatment-minus-control contrasts
jointly. For EMM order `(Control, Treatment 1, Treatment 2)`, use

$$
C=\begin{pmatrix}-1&1&0\\-1&0&1\end{pmatrix},\qquad
\hat\theta=C\hat\mu,\qquad S=CVC^\top.
$$

In particular, $S_{12}=V_{T_1T_2}-V_{T_1C}-V_{CT_2}+V_{CC}$.

Two scalar SEs lose this dependence, even for independent unadjusted arm means
because both contrasts contain the control. Model adjustment can change that
dependence further. Store the joint covariance and contrast-specific df;
separate `DatasetGeneral` analyses cannot establish familywise error control.

```{r}
#| label: multi-arm-longitudinal-extraction

multi_data <- read.csv(file.path(vignette_data_dir,
    "data_multi_arm_means_repeated_measures.csv"))
control <- "Control group"
treatments <- c("Treatment group 1", "Treatment group 2")
multi_data$group <- factor(multi_data$group, levels = c(control, treatments))
multi_stage_effects <- lapply(sort(unique(multi_data$stage)), function(k) {
    fit <- lme4::lmer(outcome ~ group + gender + covariate * visit +
        (1 + visit | subject), data = multi_data[multi_data$stage == k, ])

    # Explicit df policy; requires the corresponding optional backend.
    emm <- emmeans::emmeans(fit, "group", lmer.df = "kenward-roger", disable.pbkrtest = FALSE,
        pbkrtest.limit = Inf)
    labels <- as.character(as.data.frame(summary(emm))$group)

    stopifnot(length(labels) == 3L, setequal(labels, c(control, treatments)))

    methods <- setNames(lapply(treatments, function(trt) {
        as.numeric(labels == trt) - as.numeric(labels == control)
    }), paste(treatments, "-", control))

    effects <- emmeans::contrast(emm, method = methods, adjust = "none")
    C <- do.call(rbind, methods)

    stopifnot(isTRUE(all.equal(unname(vcov(effects)),
        unname(C %*% vcov(emm) %*% t(C)), tolerance = 1e-10)))
    stopifnot(any(grepl("kenward-roger",
        attr(summary(emm), "mesg"), ignore.case = TRUE)))
    list(model = fit, emm = emm, effects = effects, covariance = vcov(effects),
        table = as.data.frame(summary(effects, type = "link")),
        tests = as.data.frame(emmeans::test(effects, null = 0,
            side = ">", adjust = "none")))
})
```

Here `vignette_data_dir` denotes the SVN checkout's `inst/data` directory.
These are unadjusted within-stage component tests. The multi-arm analysis must
explicitly specify its family of hypotheses, intersection tests/multiplicity
procedure, stage combination weights, and any treatment-selection or dropping
rules before reporting familywise decisions or simultaneous repeated intervals.
A covariance matrix plus marginal df does not by itself specify a joint finite-df
test. The chosen engine must document its joint-distribution approximation or
other justification and validate it. Applying an `emmeans` multiplicity adjustment
and then treating those p-values as ordinary scalar stage tests is not an
implicit solution.

The vignette's factor-reordering experiment and unresolved
`directionUpper = FALSE` TODO must be replaced by label-based assertions:
permuting rows leaves each named contrast unchanged; reversing a contrast
negates its estimate and swaps/negates its interval endpoints, with the null
and test direction transformed accordingly. Changing the regression reference
level alone must not change the scientific control. Its stated continue/continue/
reject decisions for both comparisons require regeneration under the corrected
multi-arm sequential specification.

#### Two-arm repeated-measures model comparisons

The two-arm development draft makes `visit`, `baseline`, and `gender` factors
and explores random-intercept models with visit, `gender:visit`, and
`baseline:visit`, followed by simpler adjusted models. However, it repeatedly
assigns to `lmResult`, finally replacing it with `lm(outcome ~ group)` before
calling emmeans. Its displayed single-look comparison therefore tests that last
ordinary model, **not** any of the preceding mixed models. The final three-stage
bulk import also uses adjusted `lm`, not `lmer`. Those row-wise `lm` fits are
useful adapter regression checks but do not account for within-subject dependence
in longitudinal data.

The design must retain each intended model as a distinct test case. A minimal
mixed-model extraction example based on the draft is:

```{r}
#| label: two-arm-repeated-measures-extraction

repeated_data <- read.csv(file.path(vignette_data_dir,
    "data_two_arm_means_repeated_measures.csv"))
repeated_data$group <- factor(repeated_data$group,
    levels = c("Control group", "Treatment group"))
repeated_data$visit <- factor(repeated_data$visit)
repeated_data$gender <- factor(repeated_data$gender)
repeated_stage_grids <- lapply(sort(unique(repeated_data$stage)), function(k) {
    fit <- lme4::lmer(outcome ~ group + covariate + gender + visit +
        (1 | subject), data = repeated_data[repeated_data$stage == k, ])
    emm <- emmeans::emmeans(fit, "group", weights = "equal",
        lmer.df = "kenward-roger", disable.pbkrtest = FALSE,
        pbkrtest.limit = Inf)
    stopifnot(any(grepl("kenward-roger",
        attr(summary(emm), "mesg"), ignore.case = TRUE)))
    emm
})
repeated_stage_tables <- lapply(repeated_stage_grids, extract_contrast,
    treatment = "Treatment group", control = "Control group")
# A scalar extraction bridge, not validation of a sequential mixed-model design.
contrast_rows <- do.call(rbind, repeated_stage_tables)
repeated_dataset <- rpact::getDataset(est = contrast_rows$estimate,
    se = contrast_rows$SE, df = contrast_rows$df)
```

The source compares a two-sided, single-look rpact result with
`normalApproximation = TRUE` against a model coefficient's t p-value. Corrected
checks must match the null, sidedness, scale, and df method and compare directly
with the emmeans **contrast** test. A normal approximation need not equal a
finite-df t test, and a regression coefficient need not represent an averaged
EMM contrast when treatment interactions are present.

#### Shared longitudinal requirements

- Preserve the selected mixed-model df method (e.g. Kenward–Roger,
  Satterthwaite, or explicitly chosen asymptotic inference), including any
  covariance adjustment. Missing optional backends must produce an actionable
  error or an explicitly acknowledged method change; do not silently turn a
  requested finite-df analysis into asymptotic inference.
- Distinguish observation rows, unique subjects per arm, and contrast information.
  `table(group)` counts visits in these files; neither it nor reference-grid
  weights supplies independent subject counts. Keep supplied subject counts
  descriptive and calculate scalar information from the contrast SE.
- Define whether the estimand is a group difference at a specified visit or
  averaged over visits, including numeric versus factor coding of visit,
  covariate settings, and equal/proportional averaging weights. The source
  formulas have no treatment-by-visit interaction and thus specify a common
  treatment shift; an endpoint-specific treatment effect with such an interaction
  is a separate test case. Never flatten a `group | visit` grid into extra arms.
  Explicit visit selection/contrast construction must precede scalar import.
- Require independent subject cohorts across stage inputs for the independent-stage
  route. Splitting rows by `stage` is insufficient if subjects return at later
  looks. Audit subject identifiers in source data, allowing for locally reused
  IDs only with documented cohort identity. Overlapping subjects, shared nuisance
  fits, and cumulative longitudinal refits need a separately validated dependence
  model and must not silently enter this route. An `emmGrid` alone cannot verify
  the cohort contract.
- Exercise missing visits, unequal follow-up, convergence/singularity diagnostics,
  and rank-deficient fixed effects. The draft's baseline-by-visit terms may be
  redundant because baseline is defined from visit. Check estimability of the
  requested contrast rather than accepting every fit or rejecting every singular
  covariance matrix. Mixed-model t tests can be approximate even within a stage;
  sequential operating-characteristic checks must include that approximation.

## 3. Correct calculation in emmeans, without rpact

For the `emm` object above, explicitly define the direction by group label,
then let emmeans calculate the contrast and its df:

```{r}
#| label: emmeans-inference
em_table <- as.data.frame(summary(emm, type = "link"))
stopifnot(nrow(em_table) == 2L,
          setequal(as.character(em_table$group), c("Control", "Treatment")))
coef <- ifelse(em_table$group == "Treatment", 1, -1)
effect <- contrast(emm, method = list("Treatment - Control" = coef),
                   adjust = "none")
effect_table <- as.data.frame(summary(
    effect, type = "link", infer = c(TRUE, TRUE),
    null = 0, side = "two-sided", adjust = "none"
))
effect_table

# Independent check of the variance calculation, on the same scale/order.
V <- vcov(emm)
se_manual <- sqrt(drop(t(coef) %*% V %*% coef))
stopifnot(isTRUE(all.equal(se_manual, effect_table$SE, tolerance = 1e-12)))

# One prespecified upper-tail hypothesis H0: theta <= theta0.
theta0 <- 0
statistic <- (effect_table$estimate - theta0) / effect_table$SE
p_upper <- pt(statistic, df = effect_table$df, lower.tail = FALSE)

# Alternatively, let emmeans perform the test and interval calculation itself.
upper_test <- as.data.frame(emmeans::test(
    effect, null = theta0, side = ">", adjust = "none"
))
ordinary_ci <- as.data.frame(confint(
    effect, level = 0.95, side = "two-sided", adjust = "none"
))
stopifnot(isTRUE(all.equal(upper_test$p.value, p_upper, tolerance = 1e-12)))
upper_test
ordinary_ci
```

The contrast returns estimate `r effect_table$estimate`, SE `r effect_table$SE`, df `r effect_table$df`, two-sided p approximately
`r effect_table$p.value`, and an ordinary 95% CI approximately [`r ordinary_ci$lower.CL`, `r ordinary_ci$upper.CL`]. This is a
single-analysis interval, not a repeated confidence interval.

`contrast()` returns another `emmGrid`; its summary evaluates df for the
contrast rather than borrowing a marginal mean's df. A single prespecified
contrast uses `adjust = "none"`; simultaneous pairwise multiplicity adjustment
is a different issue from sequential multiplicity control. See the official
[contrast documentation](https://rvlenth.github.io/emmeans/reference/contrast.html)
and [summary/test documentation](https://rvlenth.github.io/emmeans/reference/summary.emmGrid.html).

For this identity-link `lm`, link and response scales coincide. A general
interface must not mix a response-scale summary with a link-scale covariance
matrix. Ratios, transformed responses, regridded response contrasts and Bayesian
summaries require separate contracts; they are not automatically validated by
this example. The [emmeans reference-grid interface](https://rvlenth.github.io/emmeans/reference/emmeans.html)
also allows different covariate settings and averaging weights. These define
the estimand and must be consistent across stages. With treatment–covariate
interactions, this choice directly affects the treatment contrast.

## 4. Option A: scalar bridge for comparison

Option A reduces one prespecified contrast per stage to `DatasetGeneral`.
It remains useful as a reference implementation, but is not the architecture
for the full interface: it loses joint contrast covariance and arm summaries.
Option B below still delegates contrasts and df to emmeans; the choice concerns
storage and downstream analysis, not whether to use emmeans.

```{r}
#| label: general-estimates-bridge
dataset <- rpact::getDataset(
    est = effect_table$estimate, se = effect_table$SE, df = effect_table$df)
stage_result <- rpact::getStageResults(
    rpact::getDesignGroupSequential(kMax = 1),
    dataInput = dataset, thetaH0 = 0, directionUpper = TRUE)
stopifnot(isTRUE(all.equal(stage_result$pValues[1], upper_test$p.value,
    tolerance = 1e-12)))
stage_result
```

The existing general engine evaluates scalar t/z tests and supports sequential
combination. Its inverse-variance cumulative estimates and Satterthwaite-style
df do not reconstruct a pooled mixed-model fit. Keep this bridge for extraction
checks and explicitly validated scalar analyses; do not run it independently
for multiple arms and label the results familywise adjusted.

## 5. Option B: model-means dataset and analysis engine (recommended)

### 5.1 Decisions and supported analysis contracts

Implement `DatasetModelMeans` as a distinct `Dataset` subclass. Freeze the
requested EMMs, named contrasts, covariance matrices, and contrast-specific df
at construction. The initial production scope covers identity-scale Gaussian
`lm` and `lmerMod` models with independently fitted stages, including the two
longitudinal examples above. Preserve correlated random intercepts/slopes;
no conversion to independent-arm SDs is involved.

Separate three operations: import model inference, calculate intersection tests
within each stage, then apply sequential combination and closure. Select the
intersection method explicitly for multi-arm data:

| Method | Within-stage calculation | Support contract |
|---|---|---|
| `"Bonferroni"` | `min(1, length(J) * min(p[J]))` for each intersection `J` | Initial finite-df mixed-model route. Requires valid marginal p-values, but no assumed joint t distribution; inherits approximation from KR/Satterthwaite tests. |
| `"modelMaxT"` | Upper tail of the maximum statistic using the submatrix of `cov2cor(S)` | Exact within-stage multivariate t for a conventional Gaussian `lm` with common residual variance and df; asymptotic multivariate normal for explicitly selected asymptotic mixed-model inference. |

Both methods retain full covariance in the dataset. Bonferroni supplies a
concrete finite-df route for the vignette's mixed models; `modelMaxT` uses that
covariance to gain power where its joint distribution is justified. Do not
invent a joint t distribution from unequal KR/Satterthwaite df, use their
minimum as an “exact” common df, or combine KR SEs with an unrelated covariance.
For finite-df mixed-model max-T, require a separately specified and validated
joint approximation/calibration backend before enabling it. Equal numerical df
alone do not establish a common-scale multivariate t model.

The emmeans mixed-model methods can adjust covariance as well as df, and backend
availability/options can change the method used. Capture the actual method and
check it against the requested method at grid construction. See the
[emmeans model support documentation](https://rvlenth.github.io/emmeans/articles/models.html).
An existing arbitrary `emmGrid` does not carry a complete auditable record of all
user intentions; require explicit provenance when it cannot be recovered.

### 5.2 Public constructors and R6 object contract

All names in this subsection are proposed API, not existing exported functions.
Use separate constructors for means grids and precomputed contrast grids, because
both have class `emmGrid` and must not be guessed from their row labels.

| Function | Arguments and responsibility |
|---|---|
| `getDatasetFromEmmeans()` | `...`, `treatments`, `control`, `group = "group"`, `stageIds`, `estimand`, `modelInfo`, `cohortInfo`; accept separate stage grids or one list; generate label-matched treatment-minus-control contrasts. Return `DatasetModelMeans` even for two arms. |
| `getDatasetFromEmmeansContrasts()` | `...`, `contrastIds`, `stageIds`, `estimand`, `modelInfo`, `cohortInfo`; explicit contrast-input route; preserve joint covariance from each supplied contrast grid. Arm means may be absent. Do not fabricate them. |
| `.extractModelMeansStage()` | Validate one grid and provenance; use `summary(type = "link")`, `vcov()`, and `emmeans::contrast()`; return a validated stage record. |
| `.validateModelMeansDataset()` | Validate all records jointly, stable hypothesis IDs, scale/estimand, stage order and cohort declarations. |
| `.evaluateModelContrasts()` | Given a stage, named null vector and direction vector, return t/z statistics, unadjusted p-values and ordinary intervals. |
| `.getModelIntersectionPValues()` | Given stage results and the fixed hypothesis family, return one p-value per nonempty intersection using the selected method. |
| `.combineIntersectionPValues()` | Given that matrix, design, stage weights and activity mask, calculate intersection crossing history and closed elementary decisions. No arm sample-size access. |

`estimand` records outcome/units, identity scale, treatment contrast definitions,
visit target or averaging rule, covariate reference values and weighting policy.
Distinguish a fixed reference population from stage-specific empirical averaging;
permit the latter only with an explicit justification for a common target effect.
`modelInfo` is stage-indexed: model class/formula, fit method, df method requested
and used, covariance method, package versions, convergence and singularity
messages. A future model-to-grid convenience helper can collect these directly.
It must not silently refit supplied models.

`cohortInfo` contains independent-cohort confirmation, optional globally scoped
subject IDs per stage, and separately named observation/subject counts per arm.
With IDs, reject overlap; without IDs, record that independence is user-declared
and unverified. Never “fix” genuine overlap by prefixing IDs with stage. Stage-local
ID reuse requires source metadata establishing distinct cohorts.

| R6 field | Shape and invariant |
|---|---|
| `stageIds`, `contrastIds`, `armLabels` | Ordered unique identifiers. Contrast IDs refer to scientific hypotheses, independent of row/factor order. |
| `stageData` | Private list of stage records; expose copies through getters. No public independent replacement of an estimate or covariance. |
| Stage `meanTable`, `meanCovariance` | EMM labels/estimates and named `q[k] × q[k]` matrix on the same scale; optional only in contrast-input mode. Marginal EMM df are descriptive. |
| Stage `contrastMatrix` | Named `m[k] × q[k]` coefficients, aligned to EMM rows; null in contrast-input mode if not supplied. |
| Stage `contrastTable`, `contrastCovariance` | Named estimate, SE, df for each active contrast and named `m[k] × m[k]` matrix. Require `SE^2 = diag(S)`. |
| Stage `modelInfo`, `estimand`, `cohortInfo` | Provenance and diagnostic records. No use of `extras` as sample sizes. |
| `schemaVersion`, `activeContrasts` | Serialization version and stage-by-hypothesis presence. Missing future data differs from planned arm dropping. |

Call the base `Dataset` initializer without a synthetic data frame, then initialize
stage metadata through the new subclass's methods. Add
`isDatasetModelMeans()`, `getNumberOfContrasts()`, `getStageData(stage)`,
`getContrastCovariance(stage, contrastIds)`, `getContrastResults(stage)`, and
`asDatasetGeneral(contrastId)`. The last is an explicitly scalar projection for
checks; it must not erase the original object or imply a multi-arm analysis.
Override validation, stage trimming/subsetting, deep cloning, `show()`, summary,
and data-frame export; inherited means/count methods must not operate on these
records. Contrast subsetting must use principal covariance submatrices and must
not silently redefine the original testing family after looking at results.

Validate dimensions, names, symmetry and positive semidefiniteness with a
scale-relative numerical tolerance. Allow singular EMM covariance when requested
contrasts remain estimable with positive variance. Reject non-estimable requested
contrasts, nonfinite estimates/SEs, nonpositive SEs, and missing/nonpositive df;
`Inf` is allowed only for explicitly supported asymptotic inference. Reject
unselected `by` grids, unsupported transformations, and incompatible estimands.
A visit-specific grid must first select a prespecified visit or construct the
prespecified contrast; visit rows must never become extra treatment arms.

Freeze numeric summaries for supported t/z inference: at a new null, only the
numerator changes. This permits interval inversion and RDS round trips without
serializing a live model environment. Retaining an optional contrast `emmGrid`
for diagnostics is useful, but it is not authoritative after import. New linear
contrasts require re-import through emmeans, because `S` alone cannot supply their
df. Unsupported non-t/z inference needs a separate backend contract.

### 5.3 Executable stage-record prototype

The following document-local functions implement the numeric core using ordinary
lists rather than integrating R6 dispatch prematurely. They accept already
prepared identity-scale, one-row-per-arm grids; production validation and
provenance requirements are specified above. They preserve full covariance and
check the scalar tests against emmeans at a nonzero null in both directions.

```{r}
#| label: model-means-stage-prototype
prototype_stage <- function(emm, treatments, control, group = "group") {
    means <- as.data.frame(summary(emm, type = "link"))
    stopifnot(group %in% names(means), "emmean" %in% names(means),
        is.null(attr(means, "by.vars")))
    labels <- as.character(means[[group]])
    stopifnot(!anyNA(labels), !anyDuplicated(labels),
        !anyDuplicated(treatments), !control %in% treatments,
        setequal(labels, c(control, treatments)))
    C <- t(vapply(treatments, function(trt) {
        as.numeric(labels == trt) - as.numeric(labels == control)
    }, numeric(length(labels))))
    dimnames(C) <- list(paste(treatments, "-", control), labels)
    methods <- setNames(lapply(seq_len(nrow(C)), function(j) C[j, ]), rownames(C))
    effects <- emmeans::contrast(emm, methods, adjust = "none")
    tab <- as.data.frame(summary(effects, type = "link"))
    V <- unname(vcov(emm))
    S <- unname(vcov(effects))
    dimnames(V) <- list(labels, labels)
    dimnames(S) <- list(rownames(C), rownames(C))
    stopifnot(all(is.finite(tab$estimate)), all(is.finite(tab$SE)),
        all(tab$SE > 0), !anyNA(tab$df), all(tab$df > 0),
        all(is.finite(V)), all(is.finite(S)),
        isTRUE(all.equal(S, C %*% V %*% t(C), tolerance = 1e-10)),
        isTRUE(all.equal(unname(diag(S)), tab$SE^2, tolerance = 1e-10)))
    for (M in list(V, S)) {
        tol <- 1e-10 * max(abs(M))
        stopifnot(max(abs(M - t(M))) <= tol,
            min(eigen(M, symmetric = TRUE, only.values = TRUE)$values) >= -tol)
    }
    list(meanTable = means, meanCovariance = V, contrastMatrix = C,
        contrastTable = tab, contrastCovariance = S,
        # Retained only for the prototype's independent emmeans checks.
        effects = effects)
}

prototype_test <- function(stage, null = 0, directionUpper = TRUE) {
    tab <- stage$contrastTable
    stopifnot(length(null) %in% c(1L, nrow(tab)), all(is.finite(null)),
        length(directionUpper) == 1L, is.logical(directionUpper), !is.na(directionUpper))
    statistic <- (tab$estimate - null) / tab$SE
    p <- pt(if (directionUpper) statistic else -statistic,
        df = tab$df, lower.tail = FALSE)
    data.frame(contrast = tab$contrast, estimate = tab$estimate, SE = tab$SE,
        df = tab$df, statistic = statistic, p.value = p)
}

multi_model_stages <- lapply(multi_stage_effects, function(x) {
    prototype_stage(x$emm, treatments, control)
})
repeated_model_stages <- lapply(repeated_stage_grids, prototype_stage,
    treatments = "Treatment group", control = "Control group")
for (stage in c(multi_model_stages, repeated_model_stages)) {
    for (upper in c(TRUE, FALSE)) {
        observed <- prototype_test(stage, null = 0.5, directionUpper = upper)
        expected <- as.data.frame(emmeans::test(stage$effects, null = 0.5,
            side = if (upper) ">" else "<", adjust = "none"))
        stopifnot(isTRUE(all.equal(observed$p.value, expected$p.value,
            tolerance = 1e-10)))
    }
}
knitr::kable(do.call(rbind, lapply(seq_along(multi_model_stages), function(k) {
    cbind(stage = k, prototype_test(multi_model_stages[[k]]))
})), digits = 6, row.names = FALSE)
```

The multi-arm extraction retains `r nrow(multi_model_stages[[1]]$contrastMatrix)`
contrasts at each of `r length(multi_model_stages)` stages. Its stage-1 contrast
correlation is `r round(cov2cor(multi_model_stages[[1]]$contrastCovariance)[1, 2], 6)`.
This quantity is imported from the fitted model, not reconstructed from allocation.

```{r}
#| label: cohort-and-storage-checks
cohort_summary <- function(data, source) {
    cohorts <- split(as.character(data$subject), data$stage)
    unique_ids <- lapply(cohorts, unique)
    stopifnot(!anyNA(data$subject), !anyDuplicated(unlist(unique_ids)))
    data.frame(source = source, stage = names(cohorts),
        observations = lengths(cohorts), subjects = lengths(unique_ids))
}
cohort_counts <- rbind(cohort_summary(multi_data, "multi-arm"),
    cohort_summary(repeated_data, "two-arm"))
knitr::kable(cohort_counts, row.names = FALSE)
# Permuting a grid changes its storage order, not the named scientific contrasts.
permuted_stage <- prototype_stage(multi_stage_effects[[1]]$emm[c(3, 1, 2)],
    treatments, control)
stopifnot(isTRUE(all.equal(permuted_stage$contrastCovariance,
    multi_model_stages[[1]]$contrastCovariance, tolerance = 1e-10)),
    isTRUE(all.equal(permuted_stage$contrastTable$estimate,
    multi_model_stages[[1]]$contrastTable$estimate, tolerance = 1e-10)))
frozen_stages <- lapply(multi_model_stages, function(x) {
    x$effects <- NULL
    x
})
restored_stages <- unserialize(serialize(frozen_stages, NULL))
stopifnot(identical(prototype_test(restored_stages[[1]]),
    prototype_test(multi_model_stages[[1]])))
```

The supplied CSVs pass the no-overlap check on their subject IDs. That establishes
a property of these files, not proof of independent nuisance estimation in arbitrary
user models. Fits above are separate by stage. The table also demonstrates why
observation rows cannot be used as independent subject counts.

### 5.4 Intersection tests that consume model inference

For upper-tail hypotheses $H_j:\theta_j\leq\theta_{0j}$, let
$t_j=(\hat\theta_j-\theta_{0j})/\sqrt{S_{jj}}$. For lower-tail tests, negate the
statistic. Mixed directions require transforming correlation as $D R D$ with
$D_{jj}=+1$ or $-1$ before a maximum test. The prototype uses one common direction;
production nulls/directions must be matched by contrast ID without recycling.

For `modelMaxT`, the stage intersection p-value is

$$
p_J=1-F_J\!\left(\max_{j\in J} t_j\,\mathbf{1}; R_J,\nu\right),
\qquad R=\operatorname{cov2cor}(S).
$$

Here $F_J$ is the multivariate t CDF under the supported common-variance `lm`
contract, or the multivariate normal CDF under the asymptotic contract. A singleton
must reproduce its component test. Numerical integration must return its error
estimate; production code must increase accuracy or error if the decision is
unresolved at the boundary. The existing sample-size-based Dunnett correlation
is never substituted for $R_J$.

```{r}
#| label: model-intersection-prototype
prototype_intersections <- function(stage, method = c("Bonferroni", "modelMaxT"),
                                   joint = NULL, null = 0, directionUpper = TRUE) {
    method <- match.arg(method)
    tab <- prototype_test(stage, null, directionUpper)
    m <- nrow(tab)
    sets <- unlist(lapply(seq_len(m), function(size) {
        combn(seq_len(m), size, simplify = FALSE)
    }), recursive = FALSE)
    p <- vapply(sets, function(J) {
        if (method == "Bonferroni") return(min(1, length(J) * min(tab$p.value[J])))
        # joint is an explicit prototype assertion; production derives eligibility
        # from validated model provenance, not equality of reported df alone.
        stopifnot(length(joint) == 1L, joint %in% c("lm-common-t", "asymptotic"))
        if (joint == "lm-common-t") {
            stopifnot(all(is.finite(tab$df)), max(abs(tab$df - tab$df[1])) < 1e-8)
        } else stopifnot(all(is.infinite(tab$df)))
        if (length(J) == 1L) return(tab$p.value[J])
        R <- cov2cor(stage$contrastCovariance)[J, J, drop = FALSE]
        threshold <- max(if (directionUpper) tab$statistic[J] else -tab$statistic[J])
        upper <- rep(threshold, length(J))
        probability <- if (joint == "lm-common-t") {
            mvtnorm::pmvt(upper = upper, corr = R, df = tab$df[1],
                algorithm = mvtnorm::GenzBretz(maxpts = 100000, abseps = 1e-7))
        } else {
            mvtnorm::pmvnorm(upper = upper, corr = R,
                algorithm = mvtnorm::GenzBretz(maxpts = 100000, abseps = 1e-7))
        }
        stopifnot(attr(probability, "error") < 1e-5)
        max(0, min(1, 1 - as.numeric(probability)))
    }, numeric(1))
    list(sets = sets, p.value = p)
}

# Exercise covariance-aware common-t inference on independent Gaussian lm data.
# This synthetic model is not a row-wise analysis of the longitudinal CSV.
set.seed(20260911)
max_t_data <- data.frame(group = factor(rep(c("C", "T1", "T2"), each = 20)),
    x = rnorm(60))
max_t_data$y <- 2 + as.numeric(max_t_data$group) / 4 +
    max_t_data$x + rnorm(nrow(max_t_data))
max_t_fit <- lm(y ~ group + x, data = max_t_data)
max_t_stage <- prototype_stage(emmeans(max_t_fit, "group"), c("T1", "T2"), "C")
max_t_checks <- prototype_intersections(max_t_stage, "modelMaxT", "lm-common-t")
stopifnot(isTRUE(all.equal(max_t_checks$p.value[1:2],
    prototype_test(max_t_stage)$p.value, tolerance = 1e-10)))
# Compare the full-family maximum test with emmeans' multivariate-t adjustment.
max_t_reference <- as.data.frame(emmeans::test(max_t_stage$effects,
    side = ">", adjust = "mvt"))
stopifnot(abs(tail(max_t_checks$p.value, 1) - min(max_t_reference$p.value)) < 1e-4)
knitr::kable(data.frame(intersection = c("T1", "T2", "T1 & T2"),
    p.value = max_t_checks$p.value), digits = 6)
```

The `mvt` comparison above uses a model with a justified common residual df;
it does not validate a finite-df mixed-model maximum test. emmeans documents
multivariate-t adjustment and its treatment of varying df in its
[summary/test reference](https://rvlenth.github.io/emmeans/reference/summary.emmGrid.html).
The new interface must make the selected joint method visible in output.

### 5.5 Sequential combination and closed testing

For each fixed nonempty intersection $J$, calculate $p_{Jk}$ from stage $k$.
For an inverse-normal design with prespecified positive stage weights $w_k$, use

$$
Z_{J,1:k}=\frac{\sum_{i=1}^k w_i\Phi^{-1}(1-p_{Ji})}
                  {\sqrt{\sum_{i=1}^k w_i^2}}.
$$

The design's information schedule must correspond to these weight increments.
Record whether each intersection has crossed its sequential efficacy boundary
at any look up to $k$. Reject elementary $H_j$ only when **every** intersection
containing $j$ has crossed. This is closure across hypotheses and monitoring
across looks, not per-look Holm testing of marginal sequential p-values.
For Fisher designs, reuse rpact's weighted product and design-specific boundaries
through the same intersection-p-value input contract.

Bonferroni intersection p-values can be conservative; independent, super-uniform
stage p-values with fixed positive combination weights give conservative rejection
calibration under the corresponding monotone test. Mixed-model p-values are
approximate, so simulation remains a release requirement. Covariance between
contrasts within a stage does not violate independence between stage cohorts.

```{r}
#| label: closed-inverse-normal-prototype
prototype_closed_inverse_normal <- function(stages, design,
                                            method = "Bonferroni", joint = NULL) {
    # Bounded prototype: complete family, one-sided efficacy monitoring, no
    # binding futility or adaptation; weights derive from the fixed design schedule.
    stopifnot(inherits(design, "TrialDesignInverseNormal"), design$sided == 1,
        length(stages) == design$kMax)
    labels <- stages[[1]]$contrastTable$contrast
    stopifnot(all(vapply(stages, function(x) {
        identical(x$contrastTable$contrast, labels)
    }, logical(1))))
    intersections <- lapply(stages, prototype_intersections, method = method, joint = joint)
    sets <- intersections[[1]]$sets
    p <- do.call(cbind, lapply(intersections, `[[`, "p.value"))
    weights <- sqrt(diff(c(0, design$informationRates)))
    stopifnot(all(weights > 0), all(p >= 0 & p <= 1))
    # Exact p = 1 maps to -Inf; no arbitrary clipping that increases evidence.
    scores <- qnorm(p, lower.tail = FALSE)
    if (any(is.infinite(scores) & scores > 0)) {
        stop("Prototype requires non-underflowing upper-tail p-values.")
    }
    z <- vapply(seq_along(stages), function(k) {
        drop(scores[, seq_len(k), drop = FALSE] %*% weights[seq_len(k)]) /
            sqrt(sum(weights[seq_len(k)]^2))
    }, numeric(length(sets)))
    z <- matrix(z, nrow = length(sets))
    crossing <- sweep(z, 2, design$criticalValues, `>=`)
    crossed <- t(vapply(seq_along(sets), function(i) {
        cummax(crossing[i, ]) > 0
    }, logical(length(stages))))
    rejected <- t(vapply(seq_along(labels), function(j) {
        colSums(!crossed[vapply(sets, function(J) j %in% J, logical(1)), ,
            drop = FALSE]) == 0
    }, logical(length(stages))))
    dimnames(rejected) <- list(labels, paste("stage", seq_along(stages)))
    list(intersectionPValues = p, combinedStatistics = z,
        crossedIntersections = crossed, rejected = rejected)
}

prototype_design <- getDesignInverseNormal(kMax = length(multi_model_stages),
    sided = 1, alpha = 0.025, informationRates = c(1/3, 2/3, 1))
multi_closed <- prototype_closed_inverse_normal(multi_model_stages, prototype_design)
knitr::kable(multi_closed$rejected)
# With one contrast, closure reduces to the existing scalar stage-combination path.
repeated_closed <- prototype_closed_inverse_normal(repeated_model_stages, prototype_design)
scalar_check <- getStageResults(prototype_design, dataInput = repeated_dataset)
stopifnot(isTRUE(all.equal(as.numeric(repeated_closed$combinedStatistics),
    as.numeric(scalar_check$combInverseNormal), tolerance = 1e-10)))
knitr::kable(repeated_closed$rejected)
```

These tables regenerate decisions for the explicitly chosen finite-df Bonferroni
intersection/inverse-normal specification. They do not reproduce the legacy
multi-arm default Dunnett calculation, or the two-arm draft's group-sequential
analysis. Production code must consume the actual design's weights/boundary
helpers, retain stable intersection IDs, implement numerical tail handling in
log probability space, and test all design-specific stopping rules.

The stage-1 full-family Bonferroni p-value here is
`r multi_closed$intersectionPValues[nrow(multi_closed$intersectionPValues), 1]`.
An intersection p-value equal to one maps to negative infinity and prevents
later inverse-normal rejection of that intersection. This is a real power
limitation of this conservative combination, not a numerical error to hide by
clipping. For finite-df mixed models, implement Fisher combination as well (a
factor of one does not have this absorbing effect), and document this tradeoff
when selecting the intersection/design pair. Any truncated-p inverse-normal
alternative needs separately calibrated boundaries.

For the same multi-arm mixed-model fits, the following executable alternative
uses asymptotic covariance-based max-T. It explicitly rebuilds the grids with
asymptotic inference, so it does not quietly reinterpret the KR results above.
The finite-sample accuracy of this route remains subject to simulation.

```{r}
#| label: mixed-model-asymptotic-max-t
multi_asymptotic_stages <- lapply(multi_stage_effects, function(x) {
    asymptotic_emm <- emmeans::emmeans(x$model, "group", lmer.df = "asymptotic")
    prototype_stage(asymptotic_emm, treatments, control)
})
set.seed(20260912)
multi_max_t_closed <- prototype_closed_inverse_normal(
    multi_asymptotic_stages, prototype_design,
    method = "modelMaxT", joint = "asymptotic")
knitr::kable(multi_max_t_closed$rejected)
```

Keep the original hypothesis family fixed. The first implementation rejects
missing active contrasts and adaptive arm dropping. To support dropping later,
retain earlier intersection history and implement an explicit prespecified or
conditional-error-valid activity rule. Do not reduce Bonferroni denominators or
reallocate combination weights merely because later arm data are absent. The
current vignettes keep both treatments through all stages, so this restriction
does not block their illustrated multi-arm workflow.

### 5.6 Repeated inference, cumulative summaries and planning

`StageResultsModelMeans` should expose contrast-by-stage estimates, SEs, df,
statistics and component p-values, the stage covariance matrices, and an
intersection-by-stage p-value matrix. `AnalysisResultsModelMeans` should expose
the selected method, combined statistics, crossing history, closed decisions,
repeated p-values and clearly typed intervals, with existing design/result objects
composed internally where appropriate. Preserve scientific labels in printing
and plotting; show covariance/correlation on request. Do not present synthetic
SDs or label an EMM average as an observed mean.

Implement repeated p-values by rerunning the specified boundary procedure as
alpha varies, retaining the same hypothesis family, stage weights and null.
Invert the same null-dependent tests for repeated confidence sets. Closed tests
of a joint null vector do not automatically give separate rectangular intervals:
project a validated joint confidence set, or initially offer a conservative
rectangular construction by allocating family alpha across contrasts and
inverting each scalar sequential test. The latter can use fixed Bonferroni
allocation over arms (and tails for two-sided intervals); it must be labelled as
such and may not agree with closed-test decisions. Never derive a multi-arm
repeated interval from `confint(emm)` or by fixing other unknown effects at zero.

For scalar inverse-variance summaries, define $I_{jk}=1/S_{k,jj}$ and
$\hat\theta_{j,1:K}=\sum_k I_{jk}\hat\theta_{jk}/\sum_k I_{jk}$. These are
precision-weighted summaries of independent stage contrasts. A multivariate GLS
summary, if offered for a stable common effect vector and invertible $S_k$, is

$$
\hat\theta_{1:K}=\left(\sum_k S_k^{-1}\right)^{-1}
                 \sum_k S_k^{-1}\hat\theta_k,\qquad
S_{1:K}=\left(\sum_k S_k^{-1}\right)^{-1}.
$$

These covariance formulas condition on estimated stage covariance; they do not
supply exact cumulative df. Do not automatically invert singular matrices with
a generalized inverse or use these descriptive summaries as a sequential test.
No scalar “overall information fraction” exists for arbitrary changing covariance
matrices. Inverse-normal weights are design inputs, not recalculated from observed
subject counts or realized significance.

The two-arm source also requires a group-sequential route. Implement it explicitly
as an **approximate** independent-stage scalar contrast path: project to the
general engine, use its cumulative inverse-variance estimates and
Satterthwaite-style df, and record the information schedule/maximum precision.
Test the cumulative formulas and simulate the resulting monitoring procedure on
the longitudinal model before enabling this design/model combination. It is not
a pooled longitudinal refit, and transforming cumulative t p-values to normal
scores does not establish a canonical joint distribution. Keep cumulative refits
with returning subjects and multi-arm group-sequential/conditional-Dunnett designs
outside this contract until a cross-look covariance backend is implemented.

Mean-specific sample-size recalculation, `equalVariances`, `varianceOption`, and
synthetic `stDev` inputs are inapplicable: reject conflicting explicit arguments.
Conditional power, when added, must take planned contrast effects and future
information/covariance with declared df assumptions. A subject-count target needs
a model-specific mapping to precision. Audit the existing general engine's final
confidence limits and “median-unbiased” labels before exposing them; a midpoint
of repeated confidence limits is not automatically a median-unbiased estimator.

### 5.7 Repository changes and implementation order

The current `.getStageResultsMeansMultiArm()` in
[`f_analysis_multiarm_means.R`](../R/f_analysis_multiarm_means.R) computes statistics
from means/SDs. `.performClosedCombinationTest()` in
[`f_analysis_multiarm.R`](../R/f_analysis_multiarm.R) also reconstructs Dunnett
correlation from sample sizes and supplies count-derived df. **Both layers must
be bypassed or refactored**; replacing only the import is insufficient.

| Change | Concrete implementation and completion evidence |
|---|---|
| Dataset/import | Add `class_analysis_dataset_model_means.R`, the R6 contract, both constructors and assertions. Register/export/document constructors. Add tested `emmeans` dependency in `Suggests` and actionable optional-backend errors. Pass stage extraction, malformed input, deep-clone and serialization checks. |
| Dispatch | In `f_analysis_base.R`, dataset utilities/assertions and stage-result construction, detect model-means **before** existing generic multi-arm/means routes. Add `StageResultsModelMeans` and result integration without masquerading as `DatasetMeans`. |
| Component inference | Add `f_analysis_model_means.R`: named null/direction handling, frozen t/z tests, ordinary intervals, scalar projection and provenance output. Reproduce all longitudinal stage contrasts directly against emmeans, including requested-versus-actual df checks. |
| Intersection inference | Add `f_analysis_multiarm_model_means.R`: finite-df Bonferroni and eligible covariance-based max-T; return fixed-family intersection matrices, numerical integration diagnostics and method labels. |
| Sequential core | Extract the combination/boundary/closure portion of `.performClosedCombinationTest()` into an endpoint-independent helper accepting intersection p-values. Keep legacy intersection construction as a caller. Add a model caller that never queries counts or SDs. Regress existing numeric datasets against unchanged outputs. |
| Repeated inference/output | Route alpha/null inversion through the same new helper; implement conservative rectangular intervals first with explicit labels, print/summary/plots, and errors for unsupported planning/selection. Validate coverage and interpretation. |
| Group-sequential scalar path | Add an explicitly approximate model-means projection using general-estimates cumulative inference, information metadata and operating-characteristic validation for the two-arm mixed-model example. |
| Vignette migration | Replace legacy `getDataset(emmGrid, ...)` in all three workflows, name treatment/control labels, record df/estimand/cohort specifications, regenerate result tables and narrative from executable objects. |

Deliver these as Option B milestones: (1) dataset plus all `lm`/`lmer` extractions;
(2) mixed-model finite-df closed combination and eligible max-T, including repeated
inference; (3) validated scalar group-sequential integration and migration.
Milestone 1 alone is not a complete analysis release. Introduce a targeted
deprecation message for legacy lossy EMM conversion; avoid a silent patch-release
change of its class. One-arm EMM callers need a separate migration statement,
and ordinary numeric means datasets retain their existing behavior.

## 6. Recommendation and acceptance criteria

Implement Option B as the shared representation for two-arm and multi-arm model
inference. Option A remains a scalar verification bridge. Release support by
explicit model/df/intersection/design combinations, with both development
vignettes included in the required acceptance suite. The executable prototypes
above establish extraction and a bounded closed-combination workflow; they do
not replace production validation, repeated-interval implementation or operating-
characteristic studies.

The implementation should be accepted only after these checks. For simulation,
prespecify the scenario grid, replicate count, Monte Carlo error target and
acceptance tolerance before execution. Include global and partial nulls, covariance
and visit-pattern variation, small cohort sizes, missingness consistent with the
fitted model, and alternative effects; report familywise error, interval coverage,
power and stopping probabilities with Monte Carlo uncertainty. Keep extraction
unit checks separate from these operating-characteristic studies.


1. The deterministic example reproduces emmeans estimate, SE, contrast df,
   upper/lower-tail p-values and ordinary fixed-analysis interval. Include a
   nonzero null value and reversed factor ordering to catch sign errors.
2. A group-only model agrees with its ordinary independent-arm analysis; an
   adjusted balanced/orthogonal model verifies that zero covariance still
   preserves the proper model df. Include negative and positive covariance
   examples, unequal sample sizes, and a reference-grid permutation.
3. All three vignette stages reproduce the contrast calculations above. Check
   both directions, selected sequential design weights, information rates,
   repeated intervals and decisions against the chosen sequential specification.
4. Invalid grids, absent/ambiguous labels, `by` groups, non-estimable contrasts,
   unsupported scales/models and malformed stage lists give specific errors.
   Document that subject overlap and scientific estimand consistency cannot
   always be inferred from an `emmGrid` and remain part of the input contract.
5. For the multi-arm source, test all three stage fits and both treatment-versus-
   control contrasts: reproduce emmeans estimates, full `C V C'`, contrast df and
   unadjusted component tests. Cover row permutations, control labels, sign/null
   reversal, stage-list versus separate-argument imports, and the source's `lm`
   and random-intercept alternatives. Validate the chosen multiplicity/selection
   procedure under global and partial null configurations, including familywise
   error, repeated intervals, and regenerated stopping decisions.
6. For the two-arm repeated-measures source, keep each model variant separate;
   check finite-df and explicit asymptotic contrast inference, equal/proportional
   weights, visit-specific estimands, missing visits, unbalanced subject counts,
   and estimability of redundant baseline terms. Record actual df backends and
   fitting diagnostics. Test rejection of unsupported visit grids and stage
   overlap with supplied cohort metadata; document what grids alone cannot check.
   Verify that repeated observations never become synthetic independent sample
   sizes. Do not use the draft's final overwritten `lm` as mixed-model evidence.
7. Validate finite-sample sequential operating characteristics where an
   approximation is used, and audit final-inference labels/calculations before
   advertising them for this interface. Extraction agreement is necessary but
   is not a substitute for these checks.

## Execution environment

```{r}
#| label: session-info
sessionInfo()
```
