---
title: "Correct covariance handling in the rpact–emmeans interface"
date: 2026-09-11
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  message: false
params:
  vignette_csv: null
---

Design proposal, 11 September 2026. Scope: two-arm adjusted continuous-outcome
analyses, with required extensions for multi-arm comparisons and longitudinal
mixed models from the existing development vignettes. The first implementation
step remains ordinary Gaussian linear models fitted separately to non-overlapping
stages; that step alone does not complete the use cases below. No package
implementation changes are made here.

Source baseline: rpact Git commit `8c9302e4`, package version `4.5.0.9319`.
Examples were executed with R 4.6.1 and emmeans 2.0.4, loading this checkout with
`pkgload::load_all(..., compile = FALSE)`. The installed rpact version was older
(`4.5.0.9314`) and was not used for the comparisons.


## Rendering and reproducibility

Render from the repository root with
`quarto render misc/emmeans-interface-design.qmd`. This document loads the local
rpact checkout and requires `pkgload` and `emmeans` (and an available compiled
rpact library). All example and verification code is included below.

The vignette CSV is external to this repository. Set the environment variable
`RPACT_VIGNETTE_DATA` to its path, or pass
`-P vignette_csv:/path/to/data_two_arm_adjusted_means.csv` when rendering.
The default also checks the original SVN location under the current user's home.
If the CSV is unavailable, only the vignette reproduction is skipped, with a
visible note; the self-contained example still runs.

```{r}
#| label: setup
repo_root <- normalizePath("..", mustWork = TRUE)
stopifnot(file.exists(file.path(repo_root, "DESCRIPTION")))
suppressPackageStartupMessages(library(emmeans))
document_dir <- getwd()
pkgload::load_all(repo_root, quiet = TRUE, compile = FALSE)
setwd(document_dir)
options(rpact.dataset.stdev.calc.mode = "auto")
Sys.unsetenv("RPACT_DATASET_STDEV_CALC_MODE")
vignette_csv <- params$vignette_csv
if (is.null(vignette_csv)) {
    vignette_csv <- Sys.getenv("RPACT_VIGNETTE_DATA", unset = path.expand(
        "~/svn/rpact/rpact-vignettes/inst/exclusive/qmd/data/data_two_arm_adjusted_means.csv"
    ))
}
```

The following helpers compare the legacy adapter with direct emmeans inference
and the existing general-estimates interface. Their assertions check the
contrast variance and agreement of the corrected stage tests.

```{r}
#| label: comparison-helpers
extract_contrast <- function(emm, treatment, control) {
    tab <- as.data.frame(summary(emm, type = "link"))
    stopifnot(nrow(tab) == 2L,
        setequal(as.character(tab$group), c(control, treatment)))
    coefficients <- ifelse(tab$group == treatment, 1, -1)
    effect <- contrast(emm, list("Treatment - Control" = coefficients), adjust = "none")
    result <- as.data.frame(summary(effect, type = "link", infer = c(TRUE, TRUE),
        null = 0, side = "two-sided", adjust = "none"))
    manual_se <- sqrt(drop(t(coefficients) %*% vcov(emm) %*% coefficients))
    stopifnot(isTRUE(all.equal(manual_se, result$SE, tolerance = 1e-12)))
    result
}

compare_paths <- function(emm, treatment, control) {
    contrast_result <- extract_contrast(emm, treatment, control)
    design <- getDesignGroupSequential(kMax = 1)
    legacy <- getStageResults(design,
        dataInput = suppressWarnings(getDataset(emm)))
    general <- getStageResults(design, dataInput = getDataset(
        est = contrast_result$estimate, se = contrast_result$SE, df = contrast_result$df
    ))
    correct_p <- pt(contrast_result$estimate / contrast_result$SE,
        df = contrast_result$df, lower.tail = FALSE)
    emm_table <- as.data.frame(summary(emm, type = "link"))
    effect <- contrast(emm, list("Treatment - Control" =
        ifelse(emm_table$group == treatment, 1, -1)), adjust = "none")
    direct_test <- as.data.frame(emmeans::test(effect,
        null = 0, side = ">", adjust = "none"))
    stopifnot(
        isTRUE(all.equal(direct_test$p.value, correct_p, tolerance = 1e-12)),
        isTRUE(all.equal(general$testStatistics[1],
            contrast_result$estimate / contrast_result$SE, tolerance = 1e-12)),
        isTRUE(all.equal(general$pValues[1], correct_p, tolerance = 1e-12))
    )
    data.frame(
        estimate = contrast_result$estimate,
        covariance = vcov(emm)[1, 2],
        diagonal_only_se = sqrt(sum(diag(vcov(emm)))),
        legacy_se = contrast_result$estimate / legacy$testStatistics[1],
        correct_se = contrast_result$SE,
        contrast_df = contrast_result$df,
        legacy_p = legacy$pValues[1],
        correct_p = correct_p
    )
}
```

## 1. Current interface and its assumptions

### Entry points and conversion

The vignette
`rpact-vignettes/inst/exclusive/qmd/rpact_analysis_two_arm_adjusted_means_example.qmd`
fits `lm(outcome ~ group + covariate + gender)` independently within each stage,
calls `emmeans(model, "group")`, and passes one or several resulting objects to
`getDataset()`. Thus stage 2 means new stage-2 observations, not all observations
available at the second look.

The relevant implementation is:

| Component | Current responsibility |
|---|---|
| [`getDataset()` / `.getDataset()`](../R/class_analysis_dataset.R) | Accept an `emmGrid`, several such arguments, or a single list of stage objects; dispatch before ordinary numeric dataset parsing. |
| `.getDatasetMeansModelObjectsList()` in the same file | Collect `emmGrid` arguments; warn and ignore other arguments when EMM objects are present. |
| `.getDatasetMeansFromModelsByStage()` in the same file | Extract per-group summaries, construct synthetic means data, then call `getDataset()` again on a wide data frame. |
| `.getStandardDeviationFromStandardError()` in the same file | Convert marginal SEs into quantities stored as group SDs. |
| `DatasetMeans` in the same file | R6 subclass of `Dataset`, storing stage/group means, SDs and sample sizes, and their cumulative versions. |
| [Means analysis](../R/f_analysis_base_means.R) | Apply ordinary independent-sample formulas, producing `StageResultsMeans` and design-specific analysis results. |

An `emmGrid` is an emmeans S4 object. It retains the linear functions of model
coefficients defining the EMMs, coefficient estimates and covariance, and a
degrees-of-freedom calculation mechanism. `summary()` exposes marginal EMMs,
SEs and df; `vcov()` exposes the covariance matrix of the EMM estimates on the
linear-predictor scale. Covariance is available without refitting the model.
See the [emmeans object methods](https://rvlenth.github.io/emmeans/reference/emmGrid-methods.html).

The adapter currently:

1. Requires `emmGrid` stage objects. It rejects objects with more than one entry
   in their `levels` slot. The error says that models with covariates are not
   supported, but the actual check concerns the *returned grid*: an adjusted
   model summarized over `group` alone passes, as in the vignette.
2. Determines whether the model call looks like `lm` from the first stage only.
   It warns about imprecision for `lm` and experimental support for other models.
3. Reads `emmean`, `SE`, `df` from `summary(object)`, and `sigma` and `extras`
   from `as.list(object)`. Rows of `extras` are treated as sample sizes. These
   reference-grid weights are not a general contract for independent subject
   counts, particularly for weighted or repeated-measures models.
4. Moves the first EMM level to the last rpact group position. For two arms,
   EMM row 2 becomes rpact treatment group 1 and row 1 becomes control group 2.
   This assumes that the first EMM level is the control. It does not identify
   treatment from a scientific label. Choosing the control as the regression
   reference is a convenience for this convention, not a statistical requirement
   for calculating a treatment-minus-control contrast.
5. Under the default `auto` mode for `lm`, sets
   `stDev[g] = SE(EMM[g]) * sqrt(n[g])`. The environment variable
   `RPACT_DATASET_STDEV_CALC_MODE` / option `rpact.dataset.stdev.calc.mode`
   also permits `sigma`, `norm`, and `t` modes. `sigma` uses residual model SD;
   `t` uses `SE * 2 / qt(0.975, df) * sqrt(n)`. For non-`lm` models with finite
   positive df, the latter conversion also applies outside `sigma` mode.
6. Returns `DatasetMeans`, retaining neither EMM covariance nor model/contrast
   df. There is no model-aware dataset class in this conversion path.

### Downstream assumptions

For a two-arm dataset, the default is `equalVariances = TRUE` and
`normalApproximation = FALSE`. At a single stage, the means path calculates

$$
s_p^2=\frac{(n_T-1)s_T^2+(n_C-1)s_C^2}{n_T+n_C-2},\qquad
SE_{\mathrm{pooled}}=s_p\sqrt{1/n_T+1/n_C},
$$

and uses `n_T + n_C - 2` df. Setting `equalVariances = FALSE` instead uses
`sqrt(s_T^2/n_T + s_C^2/n_C)` with Welch df. Neither path includes covariance.
With synthetic SDs and unequal group sizes, even the pooled SE need not equal
the square root of the sum of the marginal EMM variances.

Cumulative means and SDs are constructed as ordinary sample summaries. This
does not generally reconstruct either a pooled ANCOVA fit or an
inverse-variance combination of independently estimated treatment contrasts.

The reduction is appropriate for a conventional independent, homoscedastic
Gaussian `lm(y ~ group)`: the EMMs are the group sample means, their covariance
is zero, and the model df are `n_T + n_C - 2`. Covariate adjustment invalidates
this reduction in general. Additional covariates do not *always* cause nonzero
covariance: balanced/orthogonal designs can still give zero covariance. Shared
coefficients alone are not a proof of correlation—the full covariance
calculation decides. Even zero covariance does not justify discarding model df.

## 2. The statistical problem

Let the fitted coefficient vector be $\hat\beta$, and let the two rows of
$L$ define the requested EMMs. Conditional on the model design,

$$
\hat\mu=L\hat\beta,\qquad V=L\widehat{\operatorname{Var}}(\hat\beta)L^\top.
$$

In EMM order `(Control, Treatment)`, set $c=(-1,1)^\top$. The desired
treatment effect and its variance are

$$
\hat\theta=c^\top\hat\mu,\qquad
\widehat{\operatorname{Var}}(\hat\theta)=c^\top Vc
=V_{TT}+V_{CC}-2V_{TC}.
$$

Keeping marginal SEs alone loses $V_{TC}$. Positive covariance makes the
diagonal-only SE too large; negative covariance makes it too small. In ANCOVA,
adjusting imbalanced groups to a common covariate value can produce negative
covariance. The problem is therefore not necessarily conservative.

There is also an independent df problem: a full-rank ordinary linear model
uses `n - rank(X)` residual df. Mixed models may have contrast-specific df.
These cannot in general be recovered from group counts or the marginal EMM df.
Rescaling an SE by a selected t quantile does not preserve the original test
distribution across null values and significance levels.

### A small reproducible ANCOVA example

This deterministic example has six observations per arm and covariate
imbalance. Its residual vector is orthogonal to all model columns, making the
calculation especially easy to inspect.

```{r}
#| label: toy-ancova
library(emmeans)
d <- data.frame(
    group = factor(rep(c("Control", "Treatment"), each = 6),
                   levels = c("Control", "Treatment")),
    x = c(0:5, 3:8)
)
d$y <- 10 + 2 * (d$group == "Treatment") + 3 * d$x +
    rep(c(1, -1, 0, 0, -1, 1), 2)
fit <- lm(y ~ group + x, data = d)
emm <- emmeans(fit, ~ group)
summary(emm)
vcov(emm)
```

At the common covariate value 4, the control and treatment EMMs are 22 and 24.
Both marginal SEs are 0.453091, and model df are 9. The EMM covariance is

```text
               Control     Treatment
Control       0.20529101   -0.05714286
Treatment    -0.05714286    0.20529101
```

Thus the correct effect is 2 and its SE is
`sqrt(0.20529101 + 0.20529101 - 2 * (-0.05714286)) = 0.724478`.
The diagonal-only SE is 0.640767. Because the group sizes are equal, this is
also the current adapter's default pooled SE in this example.

```{r}
#| label: toy-comparison
knitr::kable(compare_paths(emm, "Treatment", "Control"), digits = 6)
```


Both the missing covariance and the wrong df affect inference. Changing
`equalVariances` does not fix this balanced-size example; changing to normal
approximation does not restore covariance either. Using residual `sigma`
also fails: it omits the additional variance from covariate imbalance.

### The existing vignette is affected

Using the QMD's own `data/data_two_arm_adjusted_means.csv`, its model formula,
and its control-first level order gives the following *stage-wise*, upper-tail
tests against zero. These are not sequentially adjusted p-values.

```{r}
#| label: vignette-comparison
#| results: asis
if (file.exists(vignette_csv)) {
    raw <- read.csv(vignette_csv)
    stage_table <- do.call(rbind, lapply(sort(unique(raw$stage)), function(k) {
        stage_data <- raw[raw$stage == k, ]
        stage_data$group <- relevel(factor(stage_data$group), "Control group")
        model <- lm(outcome ~ group + covariate + gender, data = stage_data)
        stage_emm <- emmeans(model, "group")
        cbind(stage = k, n = nobs(model),
            compare_paths(stage_emm, "Treatment group", "Control group"))
    }))
    print(knitr::kable(stage_table, digits = 6, row.names = FALSE))
} else {
    cat("Vignette reproduction skipped: CSV not found. Set `vignette_csv` or ",
        "`RPACT_VIGNETTE_DATA` to reproduce this comparison.\n")
}
```


The diagonal-only SEs are respectively 9.514934, 9.754200 and 8.237943;
the small further difference from the current SEs is due to pooling synthetic
SDs with unequal arm sizes. The larger discrepancy is the omitted covariance.
The vignette's stated stopping decisions and intervals must be regenerated
after choosing the sequential analysis specification; this table alone does
not determine those decisions.

### Additional development vignettes: required coverage

The following sources were reviewed in the `rpact-vignettes` SVN checkout,
under `inst/exclusive/dev/`:

- `rpact_multi_arm_analysis_adjusted_means_example.Rmd`, using
  `inst/data/data_multi_arm_means_repeated_measures.csv`.
- `rpact_analysis_two_arm_means_repeated_measures_example.Rmd`, using
  `inst/data/data_two_arm_means_repeated_measures.csv`.

These are existing workflows to migrate, not evidence that the current adapter
provides valid model-based sequential inference. The examples below are
non-executing extraction sketches using those CSVs; they do not add rendering
dependencies or claim reproduced stopping decisions.

#### Multi-arm adjusted means with repeated measures

Despite its filename, the multi-arm example is longitudinal. It fits
`outcome ~ group + gender + covariate * visit + (1 + visit | subject)`
separately at each of three stages, obtains `emmeans(model, "group")`, and
passes successive stage grids to `getDataset()` for `getDesignInverseNormal()`
and `getAnalysisResults()`. Its two treatment groups share a control. The
random-effects term permits intercept–slope covariance, despite the source's
wording “without interaction between random effects.” The final bulk-import
examples additionally cover group-only `lm`, adjusted `lm`, and a random-intercept
model with `covariate * gender`.

The corrected interface must form both named treatment-minus-control contrasts
jointly. For EMM order `(Control, Treatment 1, Treatment 2)`, use

$$
C=\begin{pmatrix}-1&1&0\\-1&0&1\end{pmatrix},\qquad
\hat\theta=C\hat\mu,\qquad S=CVC^\top.
$$

In particular, $S_{12}=V_{T_1T_2}-V_{T_1C}-V_{CT_2}+V_{CC}$.
Two scalar SEs lose this dependence, even for independent unadjusted arm means
because both contrasts contain the control. Model adjustment can change that
dependence further. Store the joint covariance and contrast-specific df;
separate `DatasetGeneral` analyses cannot establish familywise error control.

```{r}
#| label: multi-arm-longitudinal-extraction-sketch
#| eval: false
multi_data <- read.csv(file.path(vignette_data_dir,
    "data_multi_arm_means_repeated_measures.csv"))
control <- "Control group"
treatments <- c("Treatment group 1", "Treatment group 2")
multi_data$group <- factor(multi_data$group, levels = c(control, treatments))
multi_stage_effects <- lapply(sort(unique(multi_data$stage)), function(k) {
    fit <- lme4::lmer(outcome ~ group + gender + covariate * visit +
        (1 + visit | subject), data = multi_data[multi_data$stage == k, ])
    # Explicit df policy; requires the corresponding optional backend.
    emm <- emmeans::emmeans(fit, "group", lmer.df = "kenward-roger")
    labels <- as.character(as.data.frame(summary(emm))$group)
    stopifnot(length(labels) == 3L, setequal(labels, c(control, treatments)))
    methods <- setNames(lapply(treatments, function(trt) {
        as.numeric(labels == trt) - as.numeric(labels == control)
    }), paste(treatments, "-", control))
    effects <- emmeans::contrast(emm, method = methods, adjust = "none")
    C <- do.call(rbind, methods)
    stopifnot(isTRUE(all.equal(unname(vcov(effects)),
        unname(C %*% vcov(emm) %*% t(C)), tolerance = 1e-10)))
    list(effects = effects, covariance = vcov(effects),
        table = as.data.frame(summary(effects, type = "link")),
        tests = as.data.frame(emmeans::test(effects, null = 0,
            side = ">", adjust = "none")))
})
```

Here `vignette_data_dir` denotes the SVN checkout's `inst/data` directory.
These are unadjusted within-stage component tests. The multi-arm analysis must
explicitly specify its family of hypotheses, intersection tests/multiplicity
procedure, stage combination weights, and any treatment-selection or dropping
rules before reporting familywise decisions or simultaneous repeated intervals.
A covariance matrix plus marginal df does not by itself specify a joint finite-df
test. The chosen engine must document its joint-distribution approximation or
other justification and validate it. Applying an emmeans multiplicity adjustment
and then treating those p-values as ordinary scalar stage tests is not an
implicit solution.

The vignette's factor-reordering experiment and unresolved
`directionUpper = FALSE` TODO must be replaced by label-based assertions:
permuting rows leaves each named contrast unchanged; reversing a contrast
negates its estimate and swaps/negates its interval endpoints, with the null
and test direction transformed accordingly. Changing the regression reference
level alone must not change the scientific control. Its stated continue/continue/
reject decisions for both comparisons require regeneration under the corrected
multi-arm sequential specification.

#### Two-arm repeated-measures model comparisons

The two-arm development draft makes `visit`, `baseline`, and `gender` factors
and explores random-intercept models with visit, `gender:visit`, and
`baseline:visit`, followed by simpler adjusted models. However, it repeatedly
assigns to `lmResult`, finally replacing it with `lm(outcome ~ group)` before
calling emmeans. Its displayed single-look comparison therefore tests that last
ordinary model, **not** any of the preceding mixed models. The final three-stage
bulk import also uses adjusted `lm`, not `lmer`. Those row-wise `lm` fits are
useful adapter regression checks but do not account for within-subject dependence
in longitudinal data.

The design must retain each intended model as a distinct test case. A minimal
mixed-model extraction example based on the draft is:

```{r}
#| label: two-arm-repeated-measures-extraction-sketch
#| eval: false
repeated_data <- read.csv(file.path(vignette_data_dir,
    "data_two_arm_means_repeated_measures.csv"))
repeated_data$group <- factor(repeated_data$group,
    levels = c("Control group", "Treatment group"))
repeated_data$visit <- factor(repeated_data$visit)
repeated_data$gender <- factor(repeated_data$gender)
repeated_stage_tables <- lapply(sort(unique(repeated_data$stage)), function(k) {
    fit <- lme4::lmer(outcome ~ group + covariate + gender + visit +
        (1 | subject), data = repeated_data[repeated_data$stage == k, ])
    emm <- emmeans::emmeans(fit, "group", weights = "equal",
        lmer.df = "kenward-roger")
    extract_contrast(emm, "Treatment group", "Control group")
})
# A scalar extraction bridge, not validation of a sequential mixed-model design.
contrast_rows <- do.call(rbind, repeated_stage_tables)
repeated_dataset <- rpact::getDataset(est = contrast_rows$estimate,
    se = contrast_rows$SE, df = contrast_rows$df)
```

The source compares a two-sided, single-look rpact result with
`normalApproximation = TRUE` against a model coefficient's t p-value. Corrected
checks must match the null, sidedness, scale, and df method and compare directly
with the emmeans **contrast** test. A normal approximation need not equal a
finite-df t test, and a regression coefficient need not represent an averaged
EMM contrast when treatment interactions are present.

#### Shared longitudinal requirements

- Preserve the selected mixed-model df method (e.g. Kenward–Roger,
  Satterthwaite, or explicitly chosen asymptotic inference), including any
  covariance adjustment. Missing optional backends must produce an actionable
  error or an explicitly acknowledged method change; do not silently turn a
  requested finite-df analysis into asymptotic inference.
- Distinguish observation rows, unique subjects per arm, and contrast information.
  `table(group)` counts visits in these files; neither it nor reference-grid
  weights supplies independent subject counts. Keep supplied subject counts
  descriptive and calculate scalar information from the contrast SE.
- Define whether the estimand is a group difference at a specified visit or
  averaged over visits, including numeric versus factor coding of visit,
  covariate settings, and equal/proportional averaging weights. The source
  formulas have no treatment-by-visit interaction and thus specify a common
  treatment shift; an endpoint-specific treatment effect with such an interaction
  is a separate test case. Never flatten a `group | visit` grid into extra arms.
  Explicit visit selection/contrast construction must precede scalar import.
- Require independent subject cohorts across stage inputs for the independent-stage
  route. Splitting rows by `stage` is insufficient if subjects return at later
  looks. Audit subject identifiers in source data, allowing for locally reused
  IDs only with documented cohort identity. Overlapping subjects, shared nuisance
  fits, and cumulative longitudinal refits need a separately validated dependence
  model and must not silently enter this route. An `emmGrid` alone cannot verify
  the cohort contract.
- Exercise missing visits, unequal follow-up, convergence/singularity diagnostics,
  and rank-deficient fixed effects. The draft's baseline-by-visit terms may be
  redundant because baseline is defined from visit. Check estimability of the
  requested contrast rather than accepting every fit or rejecting every singular
  covariance matrix. Mixed-model t tests can be approximate even within a stage;
  sequential operating-characteristic checks must include that approximation.

## 3. Correct calculation in emmeans, without rpact

For the `emm` object above, explicitly define the direction by group label,
then let emmeans calculate the contrast and its df:

```{r}
#| label: emmeans-inference
em_table <- as.data.frame(summary(emm, type = "link"))
stopifnot(nrow(em_table) == 2L,
          setequal(as.character(em_table$group), c("Control", "Treatment")))
coef <- ifelse(em_table$group == "Treatment", 1, -1)
effect <- contrast(emm, method = list("Treatment - Control" = coef),
                   adjust = "none")
effect_table <- as.data.frame(summary(
    effect, type = "link", infer = c(TRUE, TRUE),
    null = 0, side = "two-sided", adjust = "none"
))
effect_table

# Independent check of the variance calculation, on the same scale/order.
V <- vcov(emm)
se_manual <- sqrt(drop(t(coef) %*% V %*% coef))
stopifnot(isTRUE(all.equal(se_manual, effect_table$SE, tolerance = 1e-12)))

# One prespecified upper-tail hypothesis H0: theta <= theta0.
theta0 <- 0
statistic <- (effect_table$estimate - theta0) / effect_table$SE
p_upper <- pt(statistic, df = effect_table$df, lower.tail = FALSE)

# Alternatively, let emmeans perform the test and interval calculation itself.
upper_test <- as.data.frame(emmeans::test(
    effect, null = theta0, side = ">", adjust = "none"
))
ordinary_ci <- as.data.frame(confint(
    effect, level = 0.95, side = "two-sided", adjust = "none"
))
stopifnot(isTRUE(all.equal(upper_test$p.value, p_upper, tolerance = 1e-12)))
upper_test
ordinary_ci
```

The contrast returns estimate 2, SE 0.724478, df 9, two-sided p approximately
0.0221, and an ordinary 95% CI approximately `[0.361, 3.639]`. This is a
single-analysis interval, not a repeated confidence interval.

`contrast()` returns another `emmGrid`; its summary evaluates df for the
contrast rather than borrowing a marginal mean's df. A single prespecified
contrast uses `adjust = "none"`; simultaneous pairwise multiplicity adjustment
is a different issue from sequential multiplicity control. See the official
[contrast documentation](https://rvlenth.github.io/emmeans/reference/contrast.html)
and [summary/test documentation](https://rvlenth.github.io/emmeans/reference/summary.emmGrid.html).

For this identity-link `lm`, link and response scales coincide. A general
interface must not mix a response-scale summary with a link-scale covariance
matrix. Ratios, transformed responses, regridded response contrasts and Bayesian
summaries require separate contracts; they are not automatically validated by
this example. The [emmeans reference-grid interface](https://rvlenth.github.io/emmeans/reference/emmeans.html)
also allows different covariate settings and averaging weights. These define
the estimand and must be consistent across stages. With treatment–covariate
interactions, this choice directly affects the treatment contrast.

## 4. Option A: delegate contrast inference to emmeans (recommended)

In this option rpact literally uses emmeans functions rather than implementing
EMM contrast algebra itself: `emmeans::contrast()` defines the treatment effect,
`emmeans::test()` supplies the stage p-value for the requested null and direction,
and `confint()` supplies an ordinary fixed-analysis interval. There are two
integration depths: reuse `DatasetGeneral` immediately, or extend the analysis
interface to consume the emmeans tests directly. They share the same contrast
adapter and differ in who evaluates the stage test.

### Reuse that already exists

The checkout already has `DatasetGeneral`, an R6 subclass of `Dataset`, created
by `getDataset(est = ..., se = ..., df = ...)`. It stores one estimate, SE and
df per stage; omitted df default to infinity. It need not represent a physical
single treatment arm: the estimate can be a two-arm contrast.

[`f_analysis_base_general.R`](../R/f_analysis_base_general.R) implements:

- `StageResultsGeneral`, with stage statistics `(estimate - thetaH0) / se` and
  t-distribution p-values using the supplied df (normal for infinite df).
- Stage information `I[k] = 1 / se[k]^2` and cumulative inverse-variance
  estimates, SEs and Satterthwaite-style df:

  $$
  \hat\theta_{1:K}=\frac{\sum I_k\hat\theta_k}{\sum I_k},\quad
  SE_{1:K}=\frac1{\sqrt{\sum I_k}},\quad
  \nu_{1:K}=\frac{(\sum I_k)^2}{\sum I_k^2/\nu_k}.
  $$

- Inverse-normal and Fisher combinations using the design's weights, and a
  group-sequential path using cumulative p-values.
- Design-specific `AnalysisResultsGroupSequential`,
  `AnalysisResultsInverseNormal` and `AnalysisResultsFisher` objects, including
  repeated-inference calculations. Mean-specific assumptions and conditional
  power/sample-size parameters are marked not applicable on this path.

The existing information-rate helper in
[`f_analysis_boundary_recalculation.R`](../R/f_analysis_boundary_recalculation.R)
uses cumulative inverse SE squared for general datasets. Thus maximum
information has precision units here, not patient-count units.

The following bridge works already, with no change to rpact:

```{r}
#| label: general-estimates-bridge
dataset <- rpact::getDataset(
    est = effect_table$estimate,
    se = effect_table$SE,
    df = effect_table$df
)
stage_result <- rpact::getStageResults(
    rpact::getDesignGroupSequential(kMax = 1),
    dataInput = dataset, thetaH0 = 0, directionUpper = TRUE
)
stopifnot(isTRUE(all.equal(stage_result$pValues[1], upper_test$p.value,
    tolerance = 1e-12)))
stage_result
```

This reproduces the emmeans one-sided statistic and p-value. For multiple
independent stages, construct one contrast table per stage and pass vectors of
their estimates, SEs and df. Do not pass cumulative model fits as stage inputs.

### Direct emmeans p-value backend

The minimal bridge above still recomputes p-values in rpact from the contrast
estimate, SE and df. For the supported scalar t/z tests, these equal the
emmeans p-values. If the intended interface is to delegate *all within-stage
inference*, extend it with a model-test backend that retains each contrast
`emmGrid` and implements an operation such as
`evaluateStage(thetaH0, directionUpper)`. Internally it would call:

```{r}
#| label: proposed-test-backend
#| eval: false
emmeans::test(
    effect,
    null = thetaH0,
    side = if (directionUpper) ">" else "<",
    adjust = "none"
)
```

This backend would return the tested null, direction, estimate, SE, df,
statistic and the *actual emmeans p-value*. A new dedicated dataset/backend
type and corresponding dispatch would be needed: `DatasetGeneral` currently
has no public field for supplied stage p-values or a null-dependent test
function. Merely adding a `pValues` argument to `getDataset()` would not provide
this functionality.

For inverse-normal and Fisher designs, refactor stage combination into a
shared helper accepting validated one-sided stage p-values and design weights.
Both `.getStageResultsGeneral()` and the emmeans backend could call it. This
keeps emmeans responsible for the model test and rpact responsible for
sequential combination, boundaries and stopping decisions. P-values must be
unadjusted for pairwise comparisons, calibrated for the prespecified hypothesis,
and valid under the design's stage-dependence assumptions.

A vector of p-values at zero is insufficient for repeated confidence intervals
or nonzero-null tests. The backend must evaluate `emmeans::test()` again at
each candidate null used by rpact's interval inversion. Cache the contrast
object, not just the p-value. Its lifetime/serialization and emmeans version
then become part of the dataset contract. Ordinary `confint(effect)` intervals
must remain clearly separate from rpact's repeated/sequential intervals.

For the group-sequential path, stage p-values alone do not define the cumulative
model test. The implementation must still specify inverse-variance combination
of independent stage estimates or a separately supported cumulative-fit
analysis. Calling `test()` on cumulative fits does not justify treating those
fits as independent stages. The sequential qualifications below apply equally
to a direct emmeans backend.

**Choice within Option A:** the estimate/SE/df bridge is sufficient for the
initial ordinary-linear-model contract. The direct test backend is the explicit
choice if keeping emmeans as the authoritative evaluator of within-stage
p-values is a requirement. It adds backend/dispatch and interval-inversion work,
but avoids duplicating emmeans test evaluation in rpact. It does not by itself
extend validation to every model supported by emmeans.

### Proposed public interface and implementation changes

Introduce a dedicated constructor, provisionally
`getDatasetFromEmmeans(..., treatment, control, group = "group")`. This is a
proposal, not an existing function. For the initial release, accept one
two-row EMM grid per stage, or a list of those grids, for supported ordinary
Gaussian linear models. Explicit arm labels remove the dependence on reference
level ordering. A separate explicitly selected contrast-input mode could later
accept one already constructed contrast per stage; since both means and
contrasts have class `emmGrid`, class alone cannot distinguish them.

The new constructor should:

1. Validate the number of rows, group labels, stage ordering, absence of `by`
   strata, supported model/scale, and consistent estimand across stages.
   Check every stage rather than using the first stage's model call.
2. Build label-matched treatment-minus-control coefficients; call
   `emmeans::contrast()` and summarize on the declared scale with no pairwise
   multiplicity adjustment. Use the returned estimate, SE and contrast df;
   with the direct backend, retain the contrast for `emmeans::test()` as well.
3. Reject non-estimable contrasts, missing/nonfinite estimates, nonpositive or
   nonfinite SEs, and invalid df; allow `Inf` df when the supported inference
   method supplies them. Never silently replace unavailable df with `Inf`.
4. Return `getDataset(est = ..., se = ..., df = ...)`. Covariance is already
   incorporated in the SE; retaining the full matrix is not necessary for
   inference on this one prespecified scalar contrast.
5. Preserve arm labels, contrast direction, scale, model/df method, and
   reference-grid settings as supported provenance metadata if a small
   extension to `DatasetGeneral` is needed. Sample sizes and arm EMMs may be
   descriptive metadata, not synthetic inputs to the means engine.

Code changes would be concentrated in the constructor and its dispatch,
documentation and tests. Add emmeans to `Suggests` with an appropriate tested
minimum version, use `requireNamespace()` with an actionable message, and
export/document the new function. emmeans is currently absent from both
`Imports` and `Suggests`. Prefer its public methods over the current
`sigma`/`extras` extraction. Do not add a new argument inside the present
`getDataset(..., treatment = ...)` convention without fixing its parser:
non-`emmGrid` arguments are currently ignored with a warning.

Deprecate the lossy `getDataset(emmGrid, ...)` conversion with a specific
migration message and update the vignette to the new constructor (or the
explicit bridge first). Avoid silently changing its result from `DatasetMeans`
to `DatasetGeneral` in a patch release: existing callers may access arm means,
SDs and sample sizes. A later documented breaking change could make the old
entry point delegate to the new contract. Ordinary numeric means datasets need
no change. Existing multi-arm and one-arm EMM calls need an explicit migration
policy rather than being routed through a two-arm-only adapter.

### Sequential inference is a separate validation requirement

Correct contrast extraction fixes the within-stage input. It does not prove
that every downstream sequential procedure is exact for every fitted model.

For independent stage data and separately fitted ordinary Gaussian linear
models, the null stage p-values are valid conditional on the model designs.
An inverse-normal or Fisher combination with appropriately prespecified weights
is a natural route. Shared participants, pooled nuisance estimation across
stages, or cumulative refits do not automatically give independent stage tests.

For group-sequential monitoring, canonical joint-distribution assumptions and
the information-time rule must also hold. Inverse-variance pooling with
estimated stage variances and the displayed Satterthwaite df is generally an
approximation, not the exact t-test from refitting ANCOVA to all observations.
Converting each cumulative t p-value to a normal score does not itself establish
the required joint distribution. A corrected group-sequential vignette needs
an explicit analysis specification and operating-characteristic validation;
it must not inherit an exactness claim from agreement at one stage.

Also audit the existing general path's final-inference outputs before claiming
full equivalence with the means path: it currently populates final confidence
limits from repeated confidence limits and labels their midpoint as a
median-unbiased estimate. These constructions are not generally sufficient to
establish stagewise-ordering confidence limits or median unbiasedness. This is
an existing downstream issue, separate from importing covariance correctly.

**Tradeoff:** this is the smallest statistically coherent change for one
two-arm contrast and retains contrast-specific df. It intentionally changes
the dataset representation and does not provide covariance among several
treatment contrasts or retain means-specific planning functionality.

## 5. Option B: introduce a covariance-aware model-means dataset

Introduce a distinct R6 class, provisionally `DatasetModelMeans`, inheriting
from `Dataset` rather than letting an unmodified means analysis consume it.
Store for each stage:

- Named EMM vector and its full covariance matrix, on one declared scale.
- Explicit contrast matrix and resulting contrast estimates, covariance and df.
- Arm/reference-grid labels, model and df method, and optional descriptive n.

For a contrast matrix $C$, calculate $C\hat\mu$ and $CVC^\top$.
Obtain df from emmeans for the requested contrasts: `V` by itself does not
encode contrast-specific df. Either freeze the supported contrasts at import,
or retain sufficient model machinery for later df evaluation; freezing avoids
making ordinary dataset use depend on a serialized live model object.

Required changes compared with today:

| Area | Change |
|---|---|
| Import | Replace synthetic SD extraction with named EMM vectors, `vcov()` and explicit emmeans contrasts. |
| Dataset model | Add stage-indexed matrix storage, validation, subsetting, cloning/serialization, getters and print/summary support. |
| Dispatch | Add model-means detection in dataset/stage/analysis utilities; prevent fallback to independent-arm formulas. |
| Two-arm inference | Project the specified contrast to `DatasetGeneral` and reuse its scalar analysis, or implement a separate engine with equivalent validation. |
| Cumulative summaries | Define model-aware summaries; do not pool EMMs and synthetic SDs as raw sample moments. |
| Multiple contrasts | Pass `C V C'` to an explicitly supported multiplicity/sequential engine; retaining covariance alone does not make existing multi-arm procedures valid for adjusted means. |
| Information and planning | Define information for the target contrast; separate descriptive subject counts from precision. |

Validation must check matrix dimensions, row/column names and order, symmetry,
finite entries and positive semidefiniteness within numerical tolerance. A
singular full covariance may still yield an estimable contrast with positive
variance; blanket positive-definiteness rejection would be too strong.

**Tradeoff:** this retains the scientific arm summaries and supports future
correlated multi-arm contrasts. It requires substantially more class,
dispatch and output work. For the current two-arm vignette, its inferential
projection is the same scalar estimate/SE/df produced by Option A, so it does
not solve an additional scalar statistical problem. For the multi-arm development
vignette, however, joint contrast storage and a compatible analysis engine are
required; this is now a concrete requirement rather than a hypothetical extension.

## 6. Recommendation and acceptance criteria

Implement Option A first, restricting advertised support to a clearly specified
model/estimand/stage contract. Use the manual `DatasetGeneral` bridge to develop
the corrected two-arm ordinary-model vignette now. Treat this as the first
milestone, not completion of the interface redesign. The multi-arm development
vignette makes joint contrast support an actual requirement: implement Option B
or an equivalent covariance-aware contrast dataset, while continuing to delegate
contrast construction and df to emmeans. A second milestone must validate scalar
mixed-model extraction; a third must supply the joint multi-arm sequential engine.
Until those milestones pass, give explicit unsupported-use errors on the new
interface and targeted migration guidance for legacy longitudinal/multi-arm calls;
do not silently fall back to synthetic SDs or advertise all emmeans models as
supported. Deprecation documentation must cover all three source vignettes.

The implementation should be accepted only after these checks:

1. The deterministic example reproduces emmeans estimate, SE, contrast df,
   upper/lower-tail p-values and ordinary fixed-analysis interval. Include a
   nonzero null value and reversed factor ordering to catch sign errors.
2. A group-only model agrees with its ordinary independent-arm analysis; an
   adjusted balanced/orthogonal model verifies that zero covariance still
   preserves the proper model df. Include negative and positive covariance
   examples, unequal sample sizes, and a reference-grid permutation.
3. All three vignette stages reproduce the contrast calculations above. Check
   both directions, selected sequential design weights, information rates,
   repeated intervals and decisions against the chosen sequential specification.
4. Invalid grids, absent/ambiguous labels, `by` groups, non-estimable contrasts,
   unsupported scales/models and malformed stage lists give specific errors.
   Document that subject overlap and scientific estimand consistency cannot
   always be inferred from an `emmGrid` and remain part of the input contract.
5. For the multi-arm source, test all three stage fits and both treatment-versus-
   control contrasts: reproduce emmeans estimates, full `C V C'`, contrast df and
   unadjusted component tests. Cover row permutations, control labels, sign/null
   reversal, stage-list versus separate-argument imports, and the source's `lm`
   and random-intercept alternatives. Validate the chosen multiplicity/selection
   procedure under global and partial null configurations, including familywise
   error, repeated intervals, and regenerated stopping decisions.
6. For the two-arm repeated-measures source, keep each model variant separate;
   check finite-df and explicit asymptotic contrast inference, equal/proportional
   weights, visit-specific estimands, missing visits, unbalanced subject counts,
   and estimability of redundant baseline terms. Record actual df backends and
   fitting diagnostics. Test rejection of unsupported visit grids and stage
   overlap with supplied cohort metadata; document what grids alone cannot check.
   Verify that repeated observations never become synthetic independent sample
   sizes. Do not use the draft's final overwritten `lm` as mixed-model evidence.
7. Validate finite-sample sequential operating characteristics where an
   approximation is used, and audit final-inference labels/calculations before
   advertising them for this interface. Extraction agreement is necessary but
   is not a substitute for these checks.

## Execution environment

```{r}
#| label: session-info
sessionInfo()
```
