---
output:
  md_document:
    variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

 <!-- badges: start -->
[![CRAN Status](https://www.r-pkg.org/badges/version/rpact)](https://cran.r-project.org/package=rpact)
[![R-CMD-check](https://github.com/rpact-com/rpact/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rpact-com/rpact/actions/workflows/R-CMD-check.yaml)
[![Total downloads](https://cranlogs.r-pkg.org/badges/grand-total/rpact?color=blue)](https://CRAN.R-project.org/package=rpact)
[![Monthly downloads](https://cranlogs.r-pkg.org/badges/rpact?color=blue)](https://CRAN.R-project.org/package=rpact)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License: LGPL v3](https://img.shields.io/badge/License-LGPL_v3-blue.svg)](https://www.gnu.org/licenses/lgpl-3.0)
[![shinyapps.io](https://img.shields.io/badge/Shiny-shinyapps.io-blue?style=flat&labelColor=white&logo=RStudio&logoColor=blue)](https://rpact.shinyapps.io/public/)
 <!-- badges: end -->

# rpact

Confirmatory Adaptive Clinical Trial Design, Simulation, and Analysis.

## Functional Range

*	Sample size and power calculation for
    +	means (continuous endpoint)
    +	rates (binary endpoint)
    +	survival trials with
        -	piecewise accrual time and intensity
        -	piecewise exponential survival time
        -	survival times that follow a Weibull distribution
*	Fixed sample design and designs with interim analysis stages
*	Simulation tool for means, rates, and survival data
    +	Assessment of adaptive sample size/event number recalculations based on
        conditional power
    +   Assessment of treatment selection strategies in multi-arm trials
*	Adaptive analysis of means, rates, and survival data
*	Adaptive designs and analysis for multi-arm trials
*	Adaptive analysis and simulation tools for enrichment design testing means,
    rates, and hazard ratios
*	Automatic boundary recalculations during the trial for analysis with alpha
    spending approach, including under- and over-running

## Installation

Install the latest CRAN release via

```{r installation, eval=FALSE, include=TRUE, echo=TRUE, results='hide'}
install.packages("rpact")
```

## Documentation

The documentation is hosted at [https://www.rpact.org](https://www.rpact.org)

## Vignettes

The vignettes are hosted at
[https://www.rpact.com/vignettes](https://www.rpact.com/vignettes)

## The rpact user group

The *rpact project* has an active user group consisting of decision-makers and
users from the pharmaceutical industry and CROs, who meet regularly
and, e.g., discuss best practices.

We invite you to be part of the *rpact user group*: benefit from know-how, shape
open source development in Pharma!

## Use on corporate computer systems

Please [contact](https://www.rpact.com/contact) us to learn how to use `rpact`
on FDA/GxP-compliant validated corporate computer systems and how to get a copy
of the formal validation documentation that is customized and licensed for
exclusive use by your company, e.g., to fulfill regulatory requirements.
The validation documentation contains the personal access data for performing
the installation qualification with `testPackage()`.

> [www.rpact.com/contact](https://www.rpact.com/contact)

# About

* **rpact** is a comprehensive validated^[The rpact validation documentation is
  available exclusively for our customers and supporting members. For more
  information visit
  [www.rpact.com/services/sla](https://www.rpact.com/services/sla)] R package
  for clinical research which
    + enables the design and analysis of confirmatory adaptive group sequential
      designs
    + is a powerful sample size calculator
    + is a free of charge open-source software licensed under
      [LGPL-3](https://cran.r-project.org/web/licenses/LGPL-3)
    + particularly, implements the methods described in the recent monograph by
      [Wassmer and Brannath (2016)](https://doi.org/10.1007%2F978-3-319-32562-0)

> For more information please visit [www.rpact.org](https://www.rpact.org)

* **RPACT** is a company which offers
    + enterprise R/Shiny software development services 
    + technical support for the
      [rpact](https://cran.r-project.org/package=rpact) package
    + consultancy and user training for scientists using R
    + validated software solutions and R package development for clinical
      research

> For more information please visit [www.rpact.com](https://www.rpact.com)
